### 2023-08-11 lab-book.Rmd
Initial commit

# Lab book

The purpose of this file is to be a playground and place for
documentation at the same time. This is were you try different things,
and write about what you have tried and why, a place to write "dirty"
code so that the rest of your project can be kept clean. See this
mostly as a reference for your future self and use it in a way that
makes sense to you.

### 2023-08-11 manuscript.Rmd
Initial commit

---
title: "Manuscript title here"
output: html_document
author: "Your name here"
bibliography: bibliography.bib
csl: vancouver.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<!-- This is an R Markdown document. Markdown is a simple formatting
syntax for authoring HTML, PDF, and MS Word documents. For more
details on using R Markdown see http://rmarkdown.rstudio.com -->

<!-- This is a comment and will not be present in the compiled
document. Try "knitting" this document by clicking the knit button up
to the left, or by running `rmarkdown::render("manuscript.Rmd")` in
the console and you should only see headings -->

<!-- Structure and most comments are from the Strengthening the
reporting of observational studies in epidemiology (STROBE) statement,
see https://www.strobe-statement.org/. For more explanations see the
paper Strengthening the Reporting of Observational Studies in
Epidemiology (STROBE):explanation and elaboration.  -->

<!-- Indicate the study’s design with a commonly used term in the
title, for example: "Associations between resuscitation interventions
and opportunities for improvement in adult trauma patients: A cohort
study" -->

Abstract
========

<!-- Provide in the abstract an informative and balanced summary of
what was done and what was found. Not more than 300 words. -->

Background
----------

Methods
-------

Results
-------

Conclusion
----------

Introduction
============

<!-- Explain the scientific background and rationale for the
investigation being reported. End by stating the aim or specific
objectives, including any prespecified hypotheses. The introduction
should not be longer than 1-1.5 pages, typically 4-5 paragraphs, each
6-8 lines long. Support every statement with a citation -->

You can cite document here like this [@exampleKey9999]. Open the file
bibliography.bib to learn more.

Methods 
=======

Study design
------------

<!-- Present key elements of study design -->

Setting
-------

<!-- Describe the setting, locations, and relevant dates, including
periods of recruitment, exposure, follow-up, and data collection -->

Participants
------------

<!-- Cohort study: Give the eligibility criteria, and the sources and
methods of selection of participants. Describe methods of
follow-up. For matched studies, give matching criteria and number of
exposed and unexposed -->

<!-- Case-control study: Give the eligibility criteria, and the
sources and + methods of case ascertainment and control
selection. Give the rationale for the choice of cases and
controls. For matched studies, give matching criteria and the number
of controls per case -->

<!-- Cross-sectional study: Give the eligibility criteria, and the
sources and methods of selection of participants -->

Variables and data sources/measurements
---------------------------------------

<!-- Clearly define all outcomes, exposures, predictors, potential
confounders, and effect modifiers. Give diagnostic criteria, if
applicable. For each variable of interest, give sources of data and
details of methods of assessment (measurement). Describe comparability
of assessment methods if there is more than one group -->

Bias 
----

<!-- Describe any efforts to address potential sources of bias -->

Study size 
----------

<!-- Explain how the study size was arrived at -->

Quantitative variables
----------------------

<!-- Explain how quantitative variables were handled in the
analyses. If applicable, describe which groupings were chosen and why
-->

Statistical methods
-------------------

<!--

(a) Describe all statistical methods, including those used to control
for confounding

(b) Describe any methods used to examine subgroups and interactions

(c) Explain how missing data were addressed

(d)

Cohort study: If applicable, explain how loss to follow-up was addressed

Case-control study:If applicable, explain how matching of cases and
controls was addressed

Cross-sectional study: If applicable, describe analytical methods
taking account of sampling strategy

(e) Describe any sensitivity analyses
-->

Results
=======

<!--

Participants
------------

(a) Report numbers of individuals at each stage of study—eg numbers
potentially eligible, examined for eligibility, confirmed eligible,
included in the study, completing follow-up, and analysed

(b) Give reasons for non-participation at each stage

(c) Consider use of a flow diagram

Descriptive data
----------------

(a) Give characteristics of study participants (eg demographic,
clinical, social) and information on exposures and potential
confounders

(b) Indicate number of participants with missing data for each
variable of interest

(c) Cohort study — Summarise follow-up time (eg, average and total
amount)

Outcome data
------------

Cohort study — Report numbers of outcome events or summary measures
over time

Case-control study — Report numbers in each exposure category, or
summary measures of exposure

Cross-sectional study — Report numbers of outcome events or summary
measures

Main results
------------

(a) Give unadjusted estimates and, if applicable, confounder-adjusted
estimates and their precision (eg, 95% confidence interval). Make
clear which confounders were adjusted for and why they were included

(b) Report category boundaries when continuous variables were
categorized

(c) If relevant, consider translating estimates of relative risk into
absolute risk for a meaningful time period

Other analyses 
--------------

Report other analyses done—eg analyses of subgroups and interactions,
and sensitivity analyses 
-->

You can include code in this document like this:

```{r main, echo=FALSE}
source("main.R") ## This "imports" the main script file of your project and run any code in it
```

You can also embed plots:

```{r plot, echo=FALSE}
plot(pressure)
```

You can also mix text and code, so called inline code, like this: `r 2+5`.

Discussion
==========

<!--

Key results
-----------
Summarise key results with reference to study objectives

Limitations
-----------

Discuss limitations of the study, taking into account sources of
potential bias or imprecision.  Discuss both direction and magnitude
of any potential bias

Interpretation
--------------

Give a cautious overall interpretation of results considering
objectives, limitations, multiplicity of analyses, results from
similar studies, and other relevant evidence

Generalisability
----------------

Discuss the generalisability (external validity) of the study results

-->

Conclusion
==========

<!-- Give a short conclusion, corresponding to the aim -->

References
==========

<!-- Do not edit by hand, references will be inserted and formatted automatically once you knit this document -->

### 2023-08-11 functions/example_function.R

@@ -1,0 +1,7 @@
+## This file shows how to write a function in R
+
+example_function <- function() {
+    ## This is a comment and will not be interpreted by R
+    x <- 1 + 2
+    return (x)
+}

### 2023-08-11 main.R

@@ -1,0 +1,26 @@
+## Welcome!
+
+## This is your project's main script file and together with
+## manuscript.Rmd it provides and entry point for you and other people
+## coming to the project. The code in this file should give an outline
+## of the different steps conducted in your study, from importing data
+## to producing results.
+
+## This file should be relatively short, and most of the heavy
+## lifting should be done by specialised functions. These functions
+## live in the folder functions/ and you create a new function using
+## create_function().
+
+## Feel free to remove this introductory text as you get started.
+
+## Source all functions (if you tick the box "Source on save" in
+## RStudio functions will be automatically sourced when you save
+## them). They all need to be sourced however when you compile your
+## manuscript file or run this file as a job, as that happens in a
+## clean R session.
+noacsr::source_all_functions()
+
+## Import data
+data <- import_data(test = TRUE)
+
+## Whatever you do next, maybe clean data?

### 2023-08-30 manuscript.Rmd
Update manuscript.Rmd

@@
 
-You can cite document here like this [@exampleKey9999]. Open the file
-bibliography.bib to learn more.
+Injury is still a major cause of death around the world, taking the lives of around 4,4 million people each year, almost 8% of all deaths[@World_Health_Organization_2021].
+In the United States, trauma is the 4th leading cause of death among the general population and the leading cause for people between the ages of 1 and 44[@AAST_2020].
+It is also an important cause hospitalization and morbitidity among all age groups, including seniors, and are responsible for an estimated 10% of all years lived with a disability globally.
 

### 2023-08-30 manuscript.Rmd
Update manuscript.Rmd

@@
 In the United States, trauma is the 4th leading cause of death among the general population and the leading cause for people between the ages of 1 and 44[@AAST_2020].
-It is also an important cause hospitalization and morbitidity among all age groups, including seniors, and are responsible for an estimated 10% of all years lived with a disability globally.
+It is also an important cause hospitalization and morbitidity among all age groups, including seniors, and are responsible for an estimated 10% of all years lived with a disability globally[@connolly_woo_lampron_perry_2018][@World_Health_Organization_2021].
 

### 2023-08-30 manuscript.Rmd
Update manuscript.Rmd

--- 2023-08-11 manuscript.Rmd
@@
 
-You can cite document here like this [@exampleKey9999]. Open the file
-bibliography.bib to learn more.
+Injury is still a major cause of death around the world, taking the lives of around 4,4 million people each year, almost 8% of all deaths [@World_Health_Organization_2021].
+In the United States, trauma is the 4th leading cause of death among the general population and the leading cause for people between the ages of 1 and 44 [@AAST_2020].
+It is also an important cause hospitalization and morbitidity among all age groups, including seniors, and are responsible for an estimated 10% of all years lived with a disability globally [@connolly_woo_lampron_perry_2018] [@World_Health_Organization_2021].
 

### 2023-09-04 manuscript.Rmd
Update manuscript.Rmd

@@
 It is also an important cause hospitalization and morbitidity among all age groups, including seniors, and are responsible for an estimated 10% of all years lived with a disability globally [@connolly_woo_lampron_perry_2018] [@World_Health_Organization_2021].
+To combat this, dedicated trauma centers exist that have been shown to lower mortality but also improve functional outcome in trauma patients [@Nirula_Brasel_2006].
+
+Trauma teams are multidisciplinary and operate in these centers.
+They consist of individuals from the specialities of emergency medicine, surgery, intensive care, nursing and support staff and are lead by a team leader.
+Their aim is to rapidly assess and stabilise the patient, prioritise their injuries and arrange for site of definitive care.
+Emergency physicians are often the personnel activating the trauma code using anatomical and physiological criterias, mechanism of injury and information obtained from emergency service personnel.
+
+Trauma patients brought to trauma centers may receive immediate or delayed treatment depending on what protocols are present at the facility and how they are being followed [@ryb2012delayed].
+In the case of a delayed trauma team activation (DTTA) it can often be attributed to insensitivity in triage protocol and failure to recognize injury patterns [@Hoyt_Hollingsworth-Fridlund_Winchell_Simons_Holbrook_Fortiage_1994].
+Studies have also shown that different factors such as age over 55, nonwhite ethnicity, blunt assault among others was linked to a higher risk of DTTA [@ryb2012delayed].
+
+This study will assess delayed treatment at the Karolinska University Hospital in Solna between the years 2012 and 2022 through the trauma registry and trauma care quality databse.
+Our objectives are to analyse how different factors are associated with delayed treatment in patients.
 
@@
 ------------
-
-<!-- Present key elements of study design -->
+This is a registry based cohort study using data from the trauma registry and trauma care quality database at the Karolinska University Hospital in Solna.
+The trauma registry includes about 12000 patients treated between 2012 and 2022.
+The trauma care quality database is a subset of the trauma registry and includes about 6000 patients selected for review between 2014 and 2022.
+The outcome is delayed treatment, as identified by the multidisciplinary review board and recorded in the trauma care quality database.
+This project will link the two databases and assess how different patient level factors, such as age, sex, mechanism of injury, and injury severity, are associated with delayed treatment using logistic regression.
+A 5% significance level and 95% confidence levels will be used.
 

### 2023-09-04 main.R
Update main.R

@@ -1,3 +1,23 @@
+## Load packages
+library(rofi)
+
+## Import data
+imported.data <- import_data()
+
+# Merge data
+data <- merge_data(imported.data)
+
+## Prepare data
+prepared.data <- prepare_data(data)
+
+## Select only variables required for the analysis
+selected.data <- select_variables(prepared.data)
+
+
+
+
+
+
 ## Welcome!
 
 ## This is your project's main script file and together with

### 2023-09-04 main.R
Update main.R

@@ -2,7 +2,7 @@
 library(rofi)
 
 ## Import data
-imported.data <- import_data()
+imported.data <- import_data(test=TRUE)
 
 # Merge data
 data <- merge_data(imported.data)

### 2023-09-04 functions/select_variables.R
commit

@@ -1,0 +1,10 @@
+#' Select Variables
+#'
+#' Select only the variables I need for my analysis 
+#' @param dataset The compled dataset.
+#' @export 
+prepare_data <- function(dataset) {
+    variables.to.include <- c("pt_Gender", "pt_age_yrs", "ed_sbp_value", "ed_rr_value", "ed_gcs_sum", "NISS")
+    prepared.dataset <- dataset[, variables.to.include]
+    return (prepared.dataset)
+}

### 2023-09-04 main.R

@@ -4,7 +4,11 @@ library(rofi)
 ## Import data
 imported.data <- import_data(test=TRUE)
 
+<<<<<<< HEAD
 # Merge data
+=======
+## Merge data
+>>>>>>> b25655e (commit)
 data <- merge_data(imported.data)
 
 ## Prepare data
@@ -18,6 +22,10 @@ selected.data <- select_variables(prepared.data)
 
 
 
+<<<<<<< HEAD
+=======
+
+>>>>>>> b25655e (commit)
 ## Welcome!
 
 ## This is your project's main script file and together with

### 2023-09-06 lab-book.Rmd
Update lab-book.Rmd

@@
 makes sense to you.
+
+inga barn med (15 under)
+delay är OFI. vilka patienter?
+traumaregister kopplat till swetrau, inte samma dock. swetrau, NISS 15, NKS ESS 9 b.l.a.
+kvalitetsregister är det konferenser sammanställer. auditfilters faller där, kallar till konferens om OFI är positivt
+create OFI, skapar binärt ja eller nej
+kolum i dataset som heter DT_treatment typ. delta time. ph prehospital. ed emergency department
+kolum är tiden till första hjälpande insats. kolum är i minuter, kom till hjälp.
+olika faktorer som höjer tid.
+DT treatment finns ej på 43% OFI, 74% av ej OFI.

### 2023-09-06 main.R
fix merge conflict

@@ -4,11 +4,8 @@ library(rofi)
 ## Import data
 imported.data <- import_data(test=TRUE)
 
-<<<<<<< HEAD
-# Merge data
-=======
+
 ## Merge data
->>>>>>> b25655e (commit)
 data <- merge_data(imported.data)
 
 ## Prepare data
@@ -22,10 +19,8 @@ selected.data <- select_variables(prepared.data)
 
 
 
-<<<<<<< HEAD
-=======
 
->>>>>>> b25655e (commit)
+
 ## Welcome!
 
 ## This is your project's main script file and together with

### 2023-09-06 functions/select_variables.R
Lekt med dataset

@@ -1,10 +1,17 @@
+library(rofi)
+
+
 #' Select Variables
 #'
 #' Select only the variables I need for my analysis 
-#' @param dataset The compled dataset.
+#' @param dataset The complete dataset.
 #' @export 
 prepare_data <- function(dataset) {
-    variables.to.include <- c("pt_Gender", "pt_age_yrs", "ed_sbp_value", "ed_rr_value", "ed_gcs_sum", "NISS")
-    prepared.dataset <- dataset[, variables.to.include]
-    return (prepared.dataset)
+  ## Import data
+  data <- rofi::import_data(test = TRUE)
+  
+  combined.datasets <- rofi::merge_data(data, test = TRUE)
+  combined.datasets$ofi <- rofi::create_ofi(combined.datasets)
+  
+  return (combined.datasets)
 }

### 2023-09-06 main.R

@@ -1,18 +1,9 @@
 ## Load packages
-library(rofi)
 
-## Import data
-imported.data <- import_data(test=TRUE)
-
-
-## Merge data
-data <- merge_data(imported.data)
+noacsr::source_all_functions()
 
 ## Prepare data
 prepared.data <- prepare_data(data)
-
-## Select only variables required for the analysis
-selected.data <- select_variables(prepared.data)
 
 
 
@@ -41,9 +32,7 @@ selected.data <- select_variables(prepared.data)
 ## them). They all need to be sourced however when you compile your
 ## manuscript file or run this file as a job, as that happens in a
 ## clean R session.
-noacsr::source_all_functions()
 
-## Import data
-data <- import_data(test = TRUE)
+
 
 ## Whatever you do next, maybe clean data?

### 2023-09-11 manuscriptlongversion.Rmd
Made new file, one for possible publication and one for thesis handin to KI.

---
title: Trauma treatment delay
subtitle: A retrospective cohort study on the factors affecting the delays in trauma care
author: Anton Wasielewski
bibliography: bibliography.bib
csl: bmcemerg.csl
---

Introduction
============
Trauma is still a major cause of death around the world, taking the lives of around 4,4 million people each year, almost 8% of all deaths [@World_Health_Organization_2021]. In the United States, trauma is the 4th leading cause of death among the general population and the leading cause for people between the ages of 1 and 44 [@AAST_2020]. It is also an important cause hospitalization and morbidity among all age groups, including seniors, and are responsible for an estimated 10% of all years lived with a disability globally [@connolly_woo_lampron_perry_2018] [@World_Health_Organization_2021]. To combat this, trauma systems and dedicated trauma centers exist that have been shown to lower mortality but also improve functional outcome in trauma patients [@Nirula_Brasel_2006].

Trauma systems are infrastructures that exist to provide and optimize care for injured patients starting with injury recognition and triage, transport to appropriate trauma center, inpatient care and outpatient follow-up. Beyond the clinical side, it works with outreach, education and advocacy, data collection through registries, research, funding, and disaster preparedness and response. A comprehensive and functioning trauma system requires strong leadership and engagement at the trauma center, regional and national level [@connolly_woo_lampron_perry_2018].

Trauma teams are multidisciplinary and operate in these centers. They consist of individuals from the specialties of emergency medicine, surgery, intensive care, nursing and support staff and are lead by a team leader. Their aim is to rapidly assess and stabilise the patient, prioritise their injuries and arrange for site of definitive care. Emergency physicians are often the personnel activating the trauma code using anatomical and physiological criteria, mechanism of injury and information obtained from emergency service personnel.

Trauma patients brought to trauma centers may receive immediate or delayed treatment depending on what protocols are present at the facility and how they are being followed [@ryb2012delayed]. Among the preventable or possibly preventable deaths in trauma patients, a group determined by the Combined Trauma Death Review Committee, delay in treatment has been identified as a major error contributing to death, found in up to 52.9% of patients in said group according to studies [@Teixeira_Inaba_Hadjizacharia_Brown_Salim_Rhee_Browder_Noguchi_Demetriades_2007]. In the case of a delayed trauma team activation (DTTA) it can often be attributed to insensitivity in triage protocol and failure to recognize injury patterns [@Hoyt_Hollingsworth-Fridlund_Winchell_Simons_Holbrook_Fortiage_1994]. Studies have also shown that different factors such as age over 55, nonwhite ethnicity, blunt assault among others was linked to a higher risk of DTTA [@ryb2012delayed].

To determine how patient level factors are associated with delayed treatment in adult trauma patients.

Methods
=======
This is a registry based cohort study using data from the trauma registry and trauma care quality database at the Karolinska University Hospital in Solna. The trauma registry includes about 12000 patients treated between 2012 and 2022. The trauma care quality database is a subset of the trauma registry and includes about 6000 patients selected for review between 2014 and 2022. The outcome is delayed treatment, as identified by the multidisciplinary review board and recorded in the trauma care quality database. This project will link the two databases and assess how different patient level factors, such as age, sex, mechanism of injury, and injury severity, are associated with delayed treatment using logistic regression. A 5% significance level and 95% confidence levels will be used.

<!-- Expand on methods, use the STROBE statement on cohort studies -->

### 2023-09-11 manuscriptlongversion.Rmd
Writing introduction longer version

@@
 
-Introduction
-============
-Trauma is still a major cause of death around the world, taking the lives of around 4,4 million people each year, almost 8% of all deaths [@World_Health_Organization_2021].
+# Introduction
+
+Physical trauma is defined as a serious injury to the body and is often divided into two groups, blunt force trauma and penetrative trauma.
+Blunt force trauma happens when an object or force strikes the body, often causing bruising, broken bones or deep cuts.
+Examples of this could be car crashes, falls or direct blows to the body.
+Penetrative trauma is when an object pierces the skin or body and creating an open wound.
+Examples of this are gunshot wounds and stab wounds [@National_Institute_of_General_Medical_Sciences_2020].
+
+Trauma is still a major cause of death around the world today, taking the lives of around 4,4 million people each year, almost 8% of all deaths [@World_Health_Organization_2021].
 In the United States, trauma is the 4th leading cause of death among the general population and the leading cause for people between the ages of 1 and 44 [@AAST_2020].
-It is also an important cause hospitalization and morbidity among all age groups, including seniors, and are responsible for an estimated 10% of all years lived with a disability globally [@connolly_woo_lampron_perry_2018] [@World_Health_Organization_2021].
+It is also an important cause hospitalization and morbidity among all age groups, including seniors, and are responsible for an estimated 10% of all years lived with a disability globally.
+This has a significant burden on social and economic level, costing countries billions of US dollars each year in healthcare and lost productivity.
+[@connolly_woo_lampron_perry_2018] [@World_Health_Organization_2021].
+
 To combat this, trauma systems and dedicated trauma centers exist that have been shown to lower mortality but also improve functional outcome in trauma patients [@Nirula_Brasel_2006].
@@
 
-Methods
-=======
+# Methods
+
 This is a registry based cohort study using data from the trauma registry and trauma care quality database at the Karolinska University Hospital in Solna.

### 2023-09-11 manuscriptlongversion.Rmd
Continuing on intro

@@
 
-Trauma is still a major cause of death around the world today, taking the lives of around 4,4 million people each year, almost 8% of all deaths [@World_Health_Organization_2021].
+Trauma is and has long been a major cause of death around the world today, taking the lives of around 4,4 million people each year, almost 8% of all deaths [@World_Health_Organization_2021].
 In the United States, trauma is the 4th leading cause of death among the general population and the leading cause for people between the ages of 1 and 44 [@AAST_2020].

### 2023-09-12 manuscriptlongversion.Rmd
introduction

@@
 
-Physical trauma is defined as a serious injury to the body and is often divided into two groups, blunt force trauma and penetrative trauma.
+Within the realm of modern medicine, trauma care stands as an important part in healthcare and is crucial to lowering mortality and morbidity in injured patients.
+The definition of medical trauma is "a set of psychological and physiological responses to pain, injury, serious illness, medical procedures and frightening treatment experiences" [@International_Society_for_Traumatic_Stress_Studies_2020].
+The one we will be focusing on in this paper is physical trauma, which is defined as a serious injury to the body.
+It is often divided into two subgroups, blunt force trauma and penetrative trauma.
 Blunt force trauma happens when an object or force strikes the body, often causing bruising, broken bones or deep cuts.
 Examples of this could be car crashes, falls or direct blows to the body.
-Penetrative trauma is when an object pierces the skin or body and creating an open wound.
+Penetrative trauma is when an object pierces the skin or body and creates an open wound.
 Examples of this are gunshot wounds and stab wounds [@National_Institute_of_General_Medical_Sciences_2020].
@@
 In the United States, trauma is the 4th leading cause of death among the general population and the leading cause for people between the ages of 1 and 44 [@AAST_2020].
-It is also an important cause hospitalization and morbidity among all age groups, including seniors, and are responsible for an estimated 10% of all years lived with a disability globally.
+It is also an important cause of hospitalization and morbidity among all age groups, including seniors, and is responsible for an estimated 10% of all years lived with a disability globally.
 This has a significant burden on social and economic level, costing countries billions of US dollars each year in healthcare and lost productivity.
 [@connolly_woo_lampron_perry_2018] [@World_Health_Organization_2021].
-
 To combat this, trauma systems and dedicated trauma centers exist that have been shown to lower mortality but also improve functional outcome in trauma patients [@Nirula_Brasel_2006].

### 2023-09-13 manuscriptlongversion.Rmd
more on introduction
Continuing introduction long

@@
 The definition of medical trauma is "a set of psychological and physiological responses to pain, injury, serious illness, medical procedures and frightening treatment experiences" [@International_Society_for_Traumatic_Stress_Studies_2020].
+As the definition suggests, it encompasses a broad spectrum of different conditions, both physical and psychological.
 The one we will be focusing on in this paper is physical trauma, which is defined as a serious injury to the body.
@@
 A comprehensive and functioning trauma system requires strong leadership and engagement at the trauma center, regional and national level [@connolly_woo_lampron_perry_2018].
+This system is crucial to provide care for trauma patients, both in reducing morbidity and mortality in this patient group.
+Some studies have shown that treating severely injured patients at one of these establishments is associated with a 15-32% decrease in mortality due to them being more capable of treating these patients [@Candefjord_Asker_Caragounis_2020].
 
@@
 
-To determine how patient level factors are associated with delayed treatment in adult trauma patients.
+In this paper our aim is to determine what different patient level factors affect the risk of receiving delayed treatment at a trauma center.
 

### 2023-09-14 manuscriptlongversion.Rmd
intro and method

@@
 # Introduction
+
+**Definition**
 
@@
 
+**Trauma statistics**
+
 Trauma is and has long been a major cause of death around the world today, taking the lives of around 4,4 million people each year, almost 8% of all deaths [@World_Health_Organization_2021].
@@
 It is also an important cause of hospitalization and morbidity among all age groups, including seniors, and is responsible for an estimated 10% of all years lived with a disability globally.
-This has a significant burden on social and economic level, costing countries billions of US dollars each year in healthcare and lost productivity.
-[@connolly_woo_lampron_perry_2018] [@World_Health_Organization_2021].
-To combat this, trauma systems and dedicated trauma centers exist that have been shown to lower mortality but also improve functional outcome in trauma patients [@Nirula_Brasel_2006].
+This has a significant burden on social and economic level, costing countries billions of US dollars each year in healthcare and lost productivity[@connolly_woo_lampron_perry_2018].
+However, this burden is not evenly distributed between or within countries.
+Many social factors such as age, sex and social status play a major role in the risk of dying from trauma, with young men with low socio-economic status being at most risk [@World_Health_Organization_2021].
+According to the WHO, about 90% of all trauma related deaths occur in low- and middle-income countries, with injury deaths rates being higher than in high income countries.
+Even within these LMIC, people of poorer socio-economic status have higher death rates in trauma-related injuries.
+Problems identified with trauma care within these countries were infrastructure, education and training, attributed to lack of funding, brain drain to HIC and lack of availability of basic amenities [@Shanthakumar_Payne_Leitch_Alfa-Wali_2021].
+In HIC where funding and governance over the healthcare system is better designed, functioning trauma systems and dedicated trauma centers exist that have been shown to lower mortality but also improve functional outcome in trauma patients [@Nirula_Brasel_2006].
 
@@
 
-This is a registry based cohort study using data from the trauma registry and trauma care quality database at the Karolinska University Hospital in Solna.
+**Study design**
+
+This is a registry based cohort study.
+
+**Setting**
+
+The data used is from the trauma registry and trauma care quality database at the Karolinska University Hospital in Solna.
 The trauma registry includes about 12000 patients treated between 2012 and 2022.
 The trauma care quality database is a subset of the trauma registry and includes about 6000 patients selected for review between 2014 and 2022.
+
+**Participants**
+
+The database only includes patients 15 and above, so patients under that age are not included.
+
+**Variables**
+
 The outcome is delayed treatment, as identified by the multidisciplinary review board and recorded in the trauma care quality database.
@@
 
-<!-- Expand on methods, use the STROBE statement on cohort studies -->
+**Data sources/measurement**
+
+The data comes from the database
+
+**Bias**
+
+Bias is not relevant in this study as the data is collected from a database and processed by a coding program.
+
+**Study size**
+
+The study size is all of the eligible patients that was treated at the hospital between 2012 and 2022.
+
+**Quantitative variables**
+
+**Statistical methods**

### 2023-09-14 manuscriptlongversion.Rmd
Introduction more

@@
 This system is crucial to provide care for trauma patients, both in reducing morbidity and mortality in this patient group.
-Some studies have shown that treating severely injured patients at one of these establishments is associated with a 15-32% decrease in mortality due to them being more capable of treating these patients [@Candefjord_Asker_Caragounis_2020].
+Earlier studies from sweden have shown that treating severely injured patients at one of these establishments is associated with a 41% lower adjusted 30-day mortality rate compared to being treated at a non trauma care center due to them being more capable of treating these patients, with patients with higher ISS scores having an even higher protective effect with \>90% for the most critically injured patients [@Candefjord_Asker_Caragounis_2020].
+Other studies have shown similar results
 

### 2023-09-14 manuscriptlongversion.Rmd
intro

@@
 Earlier studies from sweden have shown that treating severely injured patients at one of these establishments is associated with a 41% lower adjusted 30-day mortality rate compared to being treated at a non trauma care center due to them being more capable of treating these patients, with patients with higher ISS scores having an even higher protective effect with \>90% for the most critically injured patients [@Candefjord_Asker_Caragounis_2020].
-Other studies have shown similar results
+Other studies have shown similar results, showing that treating injured patients at trauma centers is associated with a 15-30% decrease in mortality [@Moran_Lecky_Bouamra_Lawrence_Edwards_Woodford_Willett_Coats_2018][@elso_Tepas_Langland-Orban_Pracht_Papa_Lottenberg_Flint_2006]
 

### 2023-09-14 manuscriptlongversion.Rmd
Introduction

--- 2023-09-14 manuscriptlongversion.Rmd
@@
 
+**Trauma systems**
+
 Trauma systems are infrastructures that exist to provide and optimize care for injured patients starting with injury recognition and triage, transport to appropriate trauma center, inpatient care and outpatient follow-up.
@@
 Earlier studies from sweden have shown that treating severely injured patients at one of these establishments is associated with a 41% lower adjusted 30-day mortality rate compared to being treated at a non trauma care center due to them being more capable of treating these patients, with patients with higher ISS scores having an even higher protective effect with \>90% for the most critically injured patients [@Candefjord_Asker_Caragounis_2020].
-Other studies have shown similar results
+Other studies have shown similar results, showing that treating injured patients at trauma centers is associated with a 15-30% decrease in mortality [@Moran_Lecky_Bouamra_Lawrence_Edwards_Woodford_Willett_Coats_2018][@elso_Tepas_Langland-Orban_Pracht_Papa_Lottenberg_Flint_2006][@MacKenzie_Rivara_Jurkovich_Nathens_Frey_Egleston_Salkever_Scharfstein_2006][@Haas_Stukel_Gomez_Zagorski_De_Mestral_Sharma_Rubenfeld_Nathens_2012].
 
 Trauma teams are multidisciplinary and operate in these centers.
-They consist of individuals from the specialties of emergency medicine, surgery, intensive care, nursing and support staff and are lead by a team leader.
+In Sweden, they consist of individuals from the specialities of surgery, intensive care, orthopedics, nursing and support staff and are lead by a team leader.
 Their aim is to rapidly assess and stabilise the patient, prioritise their injuries and arrange for site of definitive care.
-Emergency physicians are often the personnel activating the trauma code using anatomical and physiological criteria, mechanism of injury and information obtained from emergency service personnel.
+Emergency nurses are often the personnel activating the trauma code using anatomical and physiological criteria, mechanism of injury and information obtained from emergency service personnel.
+Once in the trauma room, the team works through ABCDE to treat and stabilise the patient following the protocol present at that specific hospital [@Hedberg_Falkén_Wihlke_Jansson_Hultgren_Eriksson_Beckman_Haupt_Kjörholt_Sandersjöö_et_al._2020].
 

### 2023-09-15 manuscriptlongversion.Rmd
intro expand

@@
 ---
-title: Trauma treatment delay
-subtitle: A retrospective cohort study on the factors affecting the delays in trauma care
-author: Anton Wasielewski
+title: "Trauma treatment delay"
+author: "Anton Wasielewski"
+subtitle: A retrospective cohort study on the factors affecting the delays in trauma
+  care
+output:
+  html_document:
+    df_print: paged
+  pdf_document: default
+  word_document: default
+csl: bmcemerg.csl
 bibliography: bibliography.bib
-csl: bmcemerg.csl
 ---
@@
 This system is crucial to provide care for trauma patients, both in reducing morbidity and mortality in this patient group.
-Earlier studies from sweden have shown that treating severely injured patients at one of these establishments is associated with a 41% lower adjusted 30-day mortality rate compared to being treated at a non trauma care center due to them being more capable of treating these patients, with patients with higher ISS scores having an even higher protective effect with \>90% for the most critically injured patients [@Candefjord_Asker_Caragounis_2020].
+Earlier studies from Sweden have shown that treating severely injured patients at one of these establishments is associated with a 41% lower adjusted 30-day mortality rate compared to being treated at a non trauma care center due to them being more capable of treating these patients, with patients with higher ISS scores having an even higher protective effect with \>90% for the most critically injured patients [@Candefjord_Asker_Caragounis_2020].
 Other studies have shown similar results, showing that treating injured patients at trauma centers is associated with a 15-30% decrease in mortality [@Moran_Lecky_Bouamra_Lawrence_Edwards_Woodford_Willett_Coats_2018][@elso_Tepas_Langland-Orban_Pracht_Papa_Lottenberg_Flint_2006][@MacKenzie_Rivara_Jurkovich_Nathens_Frey_Egleston_Salkever_Scharfstein_2006][@Haas_Stukel_Gomez_Zagorski_De_Mestral_Sharma_Rubenfeld_Nathens_2012].
+During the last three decades, the introduction of trauma systems has lowered the incidence of preventable death from approximately 25% to less than 5%.
+This is attributed to the improvements in care for acute brain injuries and bleeding control.
+The incidence of late death because of sepsis and multiple organ failure, possibly a result of better and earlier resuscitation [@Asensio_Trunkey_2008].
 
@@
 Trauma patients brought to trauma centers may receive immediate or delayed treatment depending on what protocols are present at the facility and how they are being followed [@ryb2012delayed].
-Among the preventable or possibly preventable deaths in trauma patients, a group determined by the Combined Trauma Death Review Committee, delay in treatment has been identified as a major error contributing to death, found in up to 52.9% of patients in said group according to studies [@Teixeira_Inaba_Hadjizacharia_Brown_Salim_Rhee_Browder_Noguchi_Demetriades_2007].
-In the case of a delayed trauma team activation (DTTA) it can often be attributed to insensitivity in triage protocol and failure to recognize injury patterns [@Hoyt_Hollingsworth-Fridlund_Winchell_Simons_Holbrook_Fortiage_1994].
-Studies have also shown that different factors such as age over 55, nonwhite ethnicity, blunt assault among others was linked to a higher risk of DTTA [@ryb2012delayed].
+Among the preventable or possibly preventable deaths in trauma patients, a group determined by the Combined Trauma Death Review Committee, delay in treatment has been identified as a major error contributing to death, found in up to 52.9% of patients in said group [@Teixeira_Inaba_Hadjizacharia_Brown_Salim_Rhee_Browder_Noguchi_Demetriades_2007].
+In the case of a DTTA it can often be attributed to insensitivity in triage protocol and failure to recognize injury patterns [@Hoyt_Hollingsworth-Fridlund_Winchell_Simons_Holbrook_Fortiage_1994].
+Studies have also shown that different factors such as age over 55, non-white ethnicity, blunt assault among others was linked to a higher risk of DTTA [@ryb2012delayed].
 

### 2023-09-15 manuscriptlongversion.Rmd
Introduction

@@
 output:
+  word_document: default
   html_document:
@@
   pdf_document: default
-  word_document: default
-csl: bmcemerg.csl
 bibliography: bibliography.bib
@@
 Earlier studies from Sweden have shown that treating severely injured patients at one of these establishments is associated with a 41% lower adjusted 30-day mortality rate compared to being treated at a non trauma care center due to them being more capable of treating these patients, with patients with higher ISS scores having an even higher protective effect with \>90% for the most critically injured patients [@Candefjord_Asker_Caragounis_2020].
-Other studies have shown similar results, showing that treating injured patients at trauma centers is associated with a 15-30% decrease in mortality [@Moran_Lecky_Bouamra_Lawrence_Edwards_Woodford_Willett_Coats_2018][@elso_Tepas_Langland-Orban_Pracht_Papa_Lottenberg_Flint_2006][@MacKenzie_Rivara_Jurkovich_Nathens_Frey_Egleston_Salkever_Scharfstein_2006][@Haas_Stukel_Gomez_Zagorski_De_Mestral_Sharma_Rubenfeld_Nathens_2012].
+Other studies have shown similar results, showing that treating injured patients at trauma centers is associated with a 15-30% decrease in mortality [@Moran_Lecky_Bouamra_Lawrence_Edwards_Woodford_Willett_Coats_2018][@Celso_Tepas_Langland-Orban_Pracht_Papa_Lottenberg_Flint_2006][@MacKenzie_Rivara_Jurkovich_Nathens_Frey_Egleston_Salkever_Scharfstein_2006][@Haas_Stukel_Gomez_Zagorski_De_Mestral_Sharma_Rubenfeld_Nathens_2012].
 During the last three decades, the introduction of trauma systems has lowered the incidence of preventable death from approximately 25% to less than 5%.
@@
 Trauma teams are multidisciplinary and operate in these centers.
+They play a pivotal role in the initial care of the trauma patient, as they are primary care givers in the critical stage of trauma.
 In Sweden, they consist of individuals from the specialities of surgery, intensive care, orthopedics, nursing and support staff and are lead by a team leader.
@@
 Emergency nurses are often the personnel activating the trauma code using anatomical and physiological criteria, mechanism of injury and information obtained from emergency service personnel.
-Once in the trauma room, the team works through ABCDE to treat and stabilise the patient following the protocol present at that specific hospital [@Hedberg_Falkén_Wihlke_Jansson_Hultgren_Eriksson_Beckman_Haupt_Kjörholt_Sandersjöö_et_al._2020].
+Once in the trauma room, the team works through ABCDE, a systematic and structured method to assess and manage critically ill patients.
+It is designed to handle the most urgent problems first, such as airways and breathing, and then go through the rest of the body in a structured way so to not miss anything.
+[@Hedberg_Falkén_Wihlke_Jansson_Hultgren_Eriksson_Beckman_Haupt_Kjörholt_Sandersjöö_et_al._2020].
+
+**Delay in treatment**
 
 Trauma patients brought to trauma centers may receive immediate or delayed treatment depending on what protocols are present at the facility and how they are being followed [@ryb2012delayed].
+A cornerstone of trauma quality improvement programs is multidisciplinary mortality and morbidity review to identify opportunities for improvement and implement corrective actions.
+This would be a meeting where doctors from different specialities sit down and go through cases where possible OFIs were identified.
+Examples of opportunities for improvement may include lack of resources and management errors.
 Among the preventable or possibly preventable deaths in trauma patients, a group determined by the Combined Trauma Death Review Committee, delay in treatment has been identified as a major error contributing to death, found in up to 52.9% of patients in said group [@Teixeira_Inaba_Hadjizacharia_Brown_Salim_Rhee_Browder_Noguchi_Demetriades_2007].
-In the case of a DTTA it can often be attributed to insensitivity in triage protocol and failure to recognize injury patterns [@Hoyt_Hollingsworth-Fridlund_Winchell_Simons_Holbrook_Fortiage_1994].
-Studies have also shown that different factors such as age over 55, non-white ethnicity, blunt assault among others was linked to a higher risk of DTTA [@ryb2012delayed].
 

### 2023-09-16 manuscriptlongversion.Rmd
intro and method

@@
 
-The database only includes patients 15 and above, so patients under that age are not included.
+The database only includes patients 15 and above, ISS \>9 and/or trauma team activation.
+Inclusion in the study further requires that the patients have been assessed for OFIs.
+Exclusion happens if a patient is missing information in one of the examined patient factors.
 
@@
 
-The data comes from the database
+The data comes from the trauma registry and trauma care quality database.
 
@@
 
-The study size is all of the eligible patients that was treated at the hospital between 2012 and 2022.
+The study size is all of the eligible patients that was treated at the hospital between 2012 and 2022. The OFI screening started in 2013, which means that the study size can't be expanded much more.
 

### 2023-09-18 manuscriptlongversion.Rmd
intro

@@
 Examples of this are gunshot wounds and stab wounds [@National_Institute_of_General_Medical_Sciences_2020].
+Differences exist between the two groups, as blunt trauma patients seem to score higher on ISS scale on arrival, require more resources and be hospitalized for a longer period [@FITCH20191201].
 
@@
 This has a significant burden on social and economic level, costing countries billions of US dollars each year in healthcare and lost productivity[@connolly_woo_lampron_perry_2018].
+Studies estimate the cost of trauma care being between 18,500\$ and 41,500\$ depending on the country [@Willenberg_Curtis_Taylor_Jan_Glass_Myburgh_2012].
+
 However, this burden is not evenly distributed between or within countries.
@@
 The incidence of late death because of sepsis and multiple organ failure, possibly a result of better and earlier resuscitation [@Asensio_Trunkey_2008].
+Functioning trauma systems saves lives in the hospital, but their work outside of the hospital is just as important.
+Data collection from injured patients, such as mechanism of injury or mortality, are essential for creating databases that can be used for research.
+In turn, that research can be used for planning injury prevention programs that target the most common injuries in the right ways, e.g.
+teen drivers, children, specific occupations etc.
+These injury prevention programs can be planned on trauma center, organizational or government level[@ACS].
 
@@
 A cornerstone of trauma quality improvement programs is multidisciplinary mortality and morbidity review to identify opportunities for improvement and implement corrective actions.
-This would be a meeting where doctors from different specialities sit down and go through cases where possible OFIs were identified.
+This would be a meeting where doctors from different specialities sit down and go through cases where possible OFIs were identified by nurses who go through the trauma care quality database.
 Examples of opportunities for improvement may include lack of resources and management errors.
+One common OFI that this paper will look further into is delayed treatment.
 Among the preventable or possibly preventable deaths in trauma patients, a group determined by the Combined Trauma Death Review Committee, delay in treatment has been identified as a major error contributing to death, found in up to 52.9% of patients in said group [@Teixeira_Inaba_Hadjizacharia_Brown_Salim_Rhee_Browder_Noguchi_Demetriades_2007].
+Although it is such a common OFI, the patient level factors associated with delayed treatment are poorly understood.
 

### 2023-09-18 manuscriptlongversion.Rmd
intro

@@
 Examples of this are gunshot wounds and stab wounds [@National_Institute_of_General_Medical_Sciences_2020].
-Differences exist between the two groups, as blunt trauma patients seem to score higher on ISS scale on arrival, require more resources and be hospitalized for a longer period [@FITCH20191201].
+Differences exist between the two groups, as blunt trauma patients tend to score higher on ISS scale on arrival and as such require more resources and be hospitalized for a longer period because of their more sever injuries[@FITCH20191201].
 
@@
 This has a significant burden on social and economic level, costing countries billions of US dollars each year in healthcare and lost productivity[@connolly_woo_lampron_perry_2018].
-Studies estimate the cost of trauma care being between 18,500\$ and 41,500\$ depending on the country [@Willenberg_Curtis_Taylor_Jan_Glass_Myburgh_2012].
-
-However, this burden is not evenly distributed between or within countries.
+Studies estimate the cost of trauma care being between 18,500\$ and 41,500\$ in HIC, depending on the country [@Willenberg_Curtis_Taylor_Jan_Glass_Myburgh_2012].
+However, as one might expect, this burden is not evenly distributed between or within countries.
 Many social factors such as age, sex and social status play a major role in the risk of dying from trauma, with young men with low socio-economic status being at most risk [@World_Health_Organization_2021].
-According to the WHO, about 90% of all trauma related deaths occur in low- and middle-income countries, with injury deaths rates being higher than in high income countries.
+It is not only patient level factors that affect how people might be affected by trauma.
+About 90% of all trauma related deaths occur in low- and middle-income countries, with death rates by trauma also being higher than in high income countries.
 Even within these LMIC, people of poorer socio-economic status have higher death rates in trauma-related injuries.
-Problems identified with trauma care within these countries were infrastructure, education and training, attributed to lack of funding, brain drain to HIC and lack of availability of basic amenities [@Shanthakumar_Payne_Leitch_Alfa-Wali_2021].
+Problems identified within these countries that contribute to these statisitics were infrastructure, education and training, attributed to lack of funding, brain drain to HIC and lack of availability of basic amenities [@Shanthakumar_Payne_Leitch_Alfa-Wali_2021].
 In HIC where funding and governance over the healthcare system is better designed, functioning trauma systems and dedicated trauma centers exist that have been shown to lower mortality but also improve functional outcome in trauma patients [@Nirula_Brasel_2006].

### 2023-09-18 manuscriptlongversion.Rmd
intro and method

@@
 
-Trauma patients brought to trauma centers may receive immediate or delayed treatment depending on what protocols are present at the facility and how they are being followed [@ryb2012delayed].
-A cornerstone of trauma quality improvement programs is multidisciplinary mortality and morbidity review to identify opportunities for improvement and implement corrective actions.
-This would be a meeting where doctors from different specialities sit down and go through cases where possible OFIs were identified by nurses who go through the trauma care quality database.
+Once treated, the patients information and timeline of what happened enters a trauma care database, depending on what hospital.
+At Nya Karolinska Universitetssjukhuset, all patients who have been treated and had ISS\>9 or had the trauma alarm activated are included and screened for OFIs.
+A cornerstone of trauma quality improvement programs is multidisciplinary mortality and morbidity review to identify OFIs and implement corrective actions.
+This refers to a meeting where doctors from different specialities sit down and go through cases where possible OFIs were identified by nurses who go through the trauma care quality database.
 Examples of opportunities for improvement may include lack of resources and management errors.
 One common OFI that this paper will look further into is delayed treatment.
-Among the preventable or possibly preventable deaths in trauma patients, a group determined by the Combined Trauma Death Review Committee, delay in treatment has been identified as a major error contributing to death, found in up to 52.9% of patients in said group [@Teixeira_Inaba_Hadjizacharia_Brown_Salim_Rhee_Browder_Noguchi_Demetriades_2007].
+Among the preventable or possibly preventable deaths in trauma patients, a subgroup of patients determined by the Combined Trauma Death Review Committee, delay in treatment has been identified as a major error contributing to death, found in up to 52.9% of patients in said group [@Teixeira_Inaba_Hadjizacharia_Brown_Salim_Rhee_Browder_Noguchi_Demetriades_2007].
 Although it is such a common OFI, the patient level factors associated with delayed treatment are poorly understood.
 
-In this paper our aim is to determine what different patient level factors affect the risk of receiving delayed treatment at a trauma center.
+The aim of this paper is to determine what different patient level factors affect the risk of receiving delayed treatment at a trauma center.
 
@@
 
-The outcome is delayed treatment, as identified by the multidisciplinary review board and recorded in the trauma care quality database.
+The outcome is delayed treatment, which is defined as more than 30 minutes passing from arrival before receiving treatment, as identified by the multidisciplinary review board and recorded in the trauma care quality database.
 This project will link the two databases and assess how different patient level factors, such as age, sex, mechanism of injury, and injury severity, are associated with delayed treatment using logistic regression.
@@
 
-The study size is all of the eligible patients that was treated at the hospital between 2012 and 2022. The OFI screening started in 2013, which means that the study size can't be expanded much more.
+The study size is all of the eligible patients that was treated at the hospital between 2012 and 2022.
+Starting 2013 only a subgroup of patients were screened for OFIs, but starting 2017 all patients that are included in the database are screened for OFIs.
 

### 2023-09-19 functions/Create_subset.R
code

@@ -1,0 +1,8 @@
+library(rofi)
+## Creating function
+Create_subset <- function(dataset)
+  ## creating subset
+  subset_data <- prepared.data[, c("Gender", "DOB", "Deceased", "ed_emerg_proc", "pt_age_yrs", "ofi")]
+ ## Exclude patients with missing variables
+subset.data <- na.omit(subset_data[, "Gender", "DOB", "Deceased", "ed_emerg_proc", "pt_age_yrs", "ofi"])
+

### 2023-09-19 functions/prepare_data.R

@@ -9,9 +9,11 @@ library(rofi)
 prepare_data <- function(dataset) {
   ## Import data
   data <- rofi::import_data(test = TRUE)
-  
+    
+  ## Create variable combined datasets by merging
   combined.datasets <- rofi::merge_data(data, test = TRUE)
+  ## Adding ofi as variable
   combined.datasets$ofi <- rofi::create_ofi(combined.datasets)
-  
+  ## output
   return (combined.datasets)
 }

### 2023-09-19 main.R

@@ -5,7 +5,8 @@ noacsr::source_all_functions()
 ## Prepare data
 prepared.data <- prepare_data(data)
 
-
+## Create subset
+subset.data <- Create_subset(data)
 
 
 

### 2023-09-19 manuscriptlongversion.Rmd
intro and code

@@
 Although it is such a common OFI, the patient level factors associated with delayed treatment are poorly understood.
+There may be several factors that correlate with receiving delayed treatment, but they have yet to be identified.
+Mapping these factors may help in identifying patients that might be at risk for receiving delayed treatment before it happens.
 
@@
 
-This is a registry based cohort study.
+This is a registry based cohort study using data from the trauma registry and trauma care quality database at the Karolinska University Hospital in Solna.
+The trauma registry includes about 12000 patients treated between 2012 and 2022.
+The trauma care quality database is a subset of the trauma registry and includes about 6000 patients selected for review between 2014 and 2022.
+The outcome is delayed treatment, as identified by the multidisciplinary review board and recorded in the trauma care quality database.
+This project will link the two databases and assess how different patient level factors, such as age, sex, mechanism of injury, and injury severity, are associated with delayed treatment using logistic regression.
+A 5% significance level and 95% confidence levels will be used.
 
@@
 
-The data used is from the trauma registry and trauma care quality database at the Karolinska University Hospital in Solna.
-The trauma registry includes about 12000 patients treated between 2012 and 2022.
-The trauma care quality database is a subset of the trauma registry and includes about 6000 patients selected for review between 2014 and 2022.
+Every year, Karolinska University hospital, which is the equivalent of a trauma level 1 hospital, admits 1500 trauma patients each year to their facility.
+To be added to the trauma registry, a patient must be over 15 years of age, had ISS\>9 and/or had trauma code activation.
+They get added to both the Karolinska Trauma registry but also the national trauma registry (SweTrau).
+The registry includes data on vital signs, injuries, interventions and patient information.
+A subset to this database, he trauma care quality database includes information relevant for mortality and morbidity conferences
 

### 2023-09-19 functions/Create_subset.R

@@ -2,7 +2,7 @@ library(rofi)
 ## Creating function
 Create_subset <- function(dataset)
   ## creating subset
-  subset_data <- prepared.data[, c("Gender", "DOB", "Deceased", "ed_emerg_proc", "pt_age_yrs", "ofi")]
+  subset.data <- prepared.data[, c("Gender", "DOB", "Deceased", "ed_emerg_proc", "pt_age_yrs", "ofi")]
  ## Exclude patients with missing variables
-subset.data <- na.omit(subset_data[, "Gender", "DOB", "Deceased", "ed_emerg_proc", "pt_age_yrs", "ofi"])
+  newsubset.data <- na.omit(subset.data[, c("Gender", "DOB", "Deceased", "ed_emerg_proc", "pt_age_yrs", "ofi")])
 

### 2023-09-20 lab-book.Rmd
Lekt med dataset
Trauma system intro
Introduction
intro and code

@@
 kvalitetsregister är det konferenser sammanställer. auditfilters faller där, kallar till konferens om OFI är positivt
-create OFI, skapar binärt ja eller nej
-kolum i dataset som heter DT_treatment typ. delta time. ph prehospital. ed emergency department
+create OFI, skapar binärt ja eller nej.
+typ. delta time. ph prehospital. ed emergency department
 kolum är tiden till första hjälpande insats. kolum är i minuter, kom till hjälp.
@@
 DT treatment finns ej på 43% OFI, 74% av ej OFI.
+dt_ed_emerg_prod är min
+
+mycket data finns på patientfaktorer o mortalitet, men ej fördröjning
+
+konferenser ovalninga globalt, resurskrävande. bara KS i sverige. nästan alla sjukhus går igenom dödsfall.
+Finns på KI 2010-2012. hjärtstopp ung man, vet ej hur man skulle göra. intresekonflikt mellan medicin o kirurg.
+gunilla wilke? anestesisköterska styrde upp mha två läkare kvalitetsförbättringsarbete. 2013 började? 
+gick först igenom dödsfall, sen utanför mortalitet. försökte inkorporera audit filters, mellan 2013-2017 audit filter för brett. valde patienter på måfå.
+gjorde om filter till de vi har nu, fick in en till sjuksyrra. liselott applicerar alla filter, typ 9 olika. om ja, går igenom pat med gunilla.
+på konferens alla som sysslar med trauma
+
+plan B, vi har all data så det ska gå
+
+tog bort
+csl: bmcemerg.csl

### 2023-09-20 manuscriptlongversion.Rmd
intro and code

@@
 
-Within the realm of modern medicine, trauma care stands as an important part in healthcare and is crucial to lowering mortality and morbidity in injured patients.
-The definition of medical trauma is "a set of psychological and physiological responses to pain, injury, serious illness, medical procedures and frightening treatment experiences" [@International_Society_for_Traumatic_Stress_Studies_2020].
+Within modern hospital care, trauma care stands as an important part in healthcare and is crucial to lowering mortality and morbidity in injured patients.
+The definition of trauma is "a set of psychological and physiological responses to pain, injury, serious illness, medical procedures and frightening treatment experiences" [@International_Society_for_Traumatic_Stress_Studies_2020].
 As the definition suggests, it encompasses a broad spectrum of different conditions, both physical and psychological.
-The one we will be focusing on in this paper is physical trauma, which is defined as a serious injury to the body.
+The one we will be focusing on in this paper is physical trauma, the clinical entity composed of physical injury and the body's associated response, hereby referred to as trauma.
+
+**Trauma statistics**
+
 It is often divided into two subgroups, blunt force trauma and penetrative trauma.
@@
 Examples of this are gunshot wounds and stab wounds [@National_Institute_of_General_Medical_Sciences_2020].
-Differences exist between the two groups, as blunt trauma patients tend to score higher on ISS scale on arrival and as such require more resources and be hospitalized for a longer period because of their more sever injuries[@FITCH20191201].
-
-**Trauma statistics**
+Differences exist between the two groups, as blunt trauma patients tend to score higher on ISS scale on arrival and as such require more resources and be hospitalized for a longer period because of their more sever injuries.
 
@@
 Other studies have shown similar results, showing that treating injured patients at trauma centers is associated with a 15-30% decrease in mortality [@Moran_Lecky_Bouamra_Lawrence_Edwards_Woodford_Willett_Coats_2018][@Celso_Tepas_Langland-Orban_Pracht_Papa_Lottenberg_Flint_2006][@MacKenzie_Rivara_Jurkovich_Nathens_Frey_Egleston_Salkever_Scharfstein_2006][@Haas_Stukel_Gomez_Zagorski_De_Mestral_Sharma_Rubenfeld_Nathens_2012].
-During the last three decades, the introduction of trauma systems has lowered the incidence of preventable death from approximately 25% to less than 5%.
+During the last three decades, the introduction of trauma systems has contributed to lowering the incidence of preventable death .
 This is attributed to the improvements in care for acute brain injuries and bleeding control.
@@
 Emergency nurses are often the personnel activating the trauma code using anatomical and physiological criteria, mechanism of injury and information obtained from emergency service personnel.
-Once in the trauma room, the team works through ABCDE, a systematic and structured method to assess and manage critically ill patients.
-It is designed to handle the most urgent problems first, such as airways and breathing, and then go through the rest of the body in a structured way so to not miss anything.
-[@Hedberg_Falkén_Wihlke_Jansson_Hultgren_Eriksson_Beckman_Haupt_Kjörholt_Sandersjöö_et_al._2020].
+Once in the trauma room, the team works systematically to assess and manage critically ill patients.
+They handle the most urgent problems first, such as airways and breathing, and then tend to less life threatening injuries before readying the patient for transport to the next site of care [@Hedberg_Falkén_Wihlke_Jansson_Hultgren_Eriksson_Beckman_Haupt_Kjörholt_Sandersjöö_et_al._2020].
 
@@
 
-Once treated, the patients information and timeline of what happened enters a trauma care database, depending on what hospital.
-At Nya Karolinska Universitetssjukhuset, all patients who have been treated and had ISS\>9 or had the trauma alarm activated are included and screened for OFIs.
+Many hospitals, mainly in HIC, have a registry database where they log the patients information and timeline of what happened.
 A cornerstone of trauma quality improvement programs is multidisciplinary mortality and morbidity review to identify OFIs and implement corrective actions.
-This refers to a meeting where doctors from different specialities sit down and go through cases where possible OFIs were identified by nurses who go through the trauma care quality database.
+This refers to a meeting where different specialities who work with trauma care sit down and go through cases where possible OFIs were identified by nurses who go through the trauma care quality database.
 Examples of opportunities for improvement may include lack of resources and management errors.
@@
 The trauma registry includes about 12000 patients treated between 2012 and 2022.
-The trauma care quality database is a subset of the trauma registry and includes about 6000 patients selected for review between 2014 and 2022.
+The trauma care quality database is a subset of the trauma registry and includes about 6000 patients selected for review between 2013 and 2022.
 The outcome is delayed treatment, as identified by the multidisciplinary review board and recorded in the trauma care quality database.

### 2023-09-20 functions/Create_subset.R

@@ -1,8 +1,13 @@
 library(rofi)
 ## Creating function
-Create_subset <- function(dataset)
-  ## creating subset
-  subset.data <- prepared.data[, c("Gender", "DOB", "Deceased", "ed_emerg_proc", "pt_age_yrs", "ofi")]
- ## Exclude patients with missing variables
-  newsubset.data <- na.omit(subset.data[, c("Gender", "DOB", "Deceased", "ed_emerg_proc", "pt_age_yrs", "ofi")])
+Create_subset <- function(dataset){
+ 
+   ## creating subset
+  subset.datas <- prepared.data[, c("Gender", "DOB", "Deceased", "ed_emerg_proc", "pt_age_yrs", "ofi")]
+  
+  ## Exclude patients with missing variables
+  newsubset.data <- na.omit(subset.datas)
+  
+  return(newsubset.data)
+}
 

### 2023-09-20 main.R

@@ -9,7 +9,7 @@ prepared.data <- prepare_data(data)
 subset.data <- Create_subset(data)
 
 
-
+print(subset.data)
 
 
 

### 2023-09-20 manuscriptlongversion.Rmd
intro

@@
 Many hospitals, mainly in HIC, have a registry database where they log the patients information and timeline of what happened.
-A cornerstone of trauma quality improvement programs is multidisciplinary mortality and morbidity review to identify OFIs and implement corrective actions.
+This is important work, as a cornerstone of trauma quality improvement programs is multidisciplinary mortality and morbidity review to identify OFIs and implement corrective actions.
 This refers to a meeting where different specialities who work with trauma care sit down and go through cases where possible OFIs were identified by nurses who go through the trauma care quality database.
+These MoM conferences are very resource and time craving, which is why many hospitals doesn't have the means to perform them.
+In Sweden
+
 Examples of opportunities for improvement may include lack of resources and management errors.

### 2023-09-20 manuscriptlongversion.Rmd
intro and method

@@
 These MoM conferences are very resource and time craving, which is why many hospitals doesn't have the means to perform them.
-In Sweden
+In Sweden, most hospitals have mortality conferences on deceased patient cases, but only one has MoM conferences to better their trauma care on patients who lived.
 
@@
 The registry includes data on vital signs, injuries, interventions and patient information.
-A subset to this database, he trauma care quality database includes information relevant for mortality and morbidity conferences
+A subset to this database, the trauma care quality database includes information relevant for MoM conferences.
+At Karolinska, patients who die during hospital stay gets their case sent to a MoM everytime.
+However, for the rest of the patients, it is determined by two specialized nurses.
+Several audit filters are set up to flag for patients with possible OFIs.
+Flagged patients are then reviewed by these two nurses who decide if there is a possible OFI, in which case their case is sent to a MoM conference for further review.
 

### 2023-09-20 manuscriptlongversion.Rmd
method and code

@@
 This is important work, as a cornerstone of trauma quality improvement programs is multidisciplinary mortality and morbidity review to identify OFIs and implement corrective actions.
-This refers to a meeting where different specialities who work with trauma care sit down and go through cases where possible OFIs were identified by nurses who go through the trauma care quality database.
+This refers to a meeting where different specialities who work with trauma care sit down and go through cases where possible OFIs were identified.
 These MoM conferences are very resource and time craving, which is why many hospitals doesn't have the means to perform them.
 In Sweden, most hospitals have mortality conferences on deceased patient cases, but only one has MoM conferences to better their trauma care on patients who lived.
-
 Examples of opportunities for improvement may include lack of resources and management errors.
@@
 **Statistical methods**
+
+Multivariable and potentially univariable logistical regression

### 2023-09-20 functions/Create_subset.R

@@ -3,7 +3,7 @@ library(rofi)
 Create_subset <- function(dataset){
  
    ## creating subset
-  subset.datas <- prepared.data[, c("Gender", "DOB", "Deceased", "ed_emerg_proc", "pt_age_yrs", "ofi")]
+  subset.datas <- prepared.data[, c("Gender", "DOB", "Deceased", "ed_emerg_proc", "pt_age_yrs", "ofi", "Problemomrade_.FMP")]
   
   ## Exclude patients with missing variables
   newsubset.data <- na.omit(subset.datas)

### 2023-09-20 manuscriptlongversion.Rmd
Comment on progress report

--- 2023-09-20 manuscriptlongversion.Rmd
@@
 
-Within modern hospital care, trauma care stands as an important part in healthcare and is crucial to lowering mortality and morbidity in injured patients.
+Within modern hospital care, trauma care stands as an important part in healthcare and is crucial to lowering mortality and morbidity in injured patients<!-- Not only hospital trauma care, you may want to mention the entire system already here-->.
 The definition of trauma is "a set of psychological and physiological responses to pain, injury, serious illness, medical procedures and frightening treatment experiences" [@International_Society_for_Traumatic_Stress_Studies_2020].
-As the definition suggests, it encompasses a broad spectrum of different conditions, both physical and psychological.
+As the definition suggests, it encompasses a broad spectrum of different conditions, both physical and psychological<!-- I don't think you need this definition here, it doesn't really fit-->.
 The one we will be focusing on in this paper is physical trauma, the clinical entity composed of physical injury and the body's associated response, hereby referred to as trauma.
@@
 
-It is often divided into two subgroups, blunt force trauma and penetrative trauma.
-Blunt force trauma happens when an object or force strikes the body, often causing bruising, broken bones or deep cuts.
+It is often divided into two subgroups<!-- By "it", do you mean trauma?
+Suggest you just write so-->, blunt force trauma and penetrating trauma.
+Blunt force trauma is when an object or force strikes the body, often causing bruising, broken bones or deep cuts.
 Examples of this could be car crashes, falls or direct blows to the body.
-Penetrative trauma is when an object pierces the skin or body and creates an open wound.
-Examples of this are gunshot wounds and stab wounds [@National_Institute_of_General_Medical_Sciences_2020].
-Differences exist between the two groups, as blunt trauma patients tend to score higher on ISS scale on arrival and as such require more resources and be hospitalized for a longer period because of their more sever injuries.
+Penetrating<!-- More common to refer to it as penetrating, rather than penetrative--> trauma is when an object pierces the skin or body and creates an open wound.
+Examples of this<!-- By "this", do you mean penetrating trauma mechanisms?--> are gunshot wounds and stab wounds [@National_Institute_of_General_Medical_Sciences_2020].
+Differences exist between the two groups, as blunt trauma patients tend to score higher on ISS scale<!-- You haven't mentioned ISS before here.
+Do you mean that blunt trauma patients are more severely injured?--> on arrival and as such require more resources and be hospitalized for a longer period because of their more severe injuries.
 
@@
 This has a significant burden on social and economic level, costing countries billions of US dollars each year in healthcare and lost productivity[@connolly_woo_lampron_perry_2018].
-Studies estimate the cost of trauma care being between 18,500\$ and 41,500\$ in HIC, depending on the country [@Willenberg_Curtis_Taylor_Jan_Glass_Myburgh_2012].
-However, as one might expect, this burden is not evenly distributed between or within countries.
+Studies estimate the cost of trauma care being between 18,500\$ and 41,500\$ in HIC<!-- Spell out the first time you mention HIC-->, depending on the country [@Willenberg_Curtis_Taylor_Jan_Glass_Myburgh_2012].
+However, as one might expect<!-- Phrases like "as one might expect", "it is known that", "as has been shown" are usually unecessary and don't add much.
+Only use them if you need the extra words -->, this burden is not evenly distributed between or within countries.
 Many social factors such as age, sex and social status play a major role in the risk of dying from trauma, with young men with low socio-economic status being at most risk [@World_Health_Organization_2021].
-It is not only patient level factors that affect how people might be affected by trauma.
-About 90% of all trauma related deaths occur in low- and middle-income countries, with death rates by trauma also being higher than in high income countries.
-Even within these LMIC, people of poorer socio-economic status have higher death rates in trauma-related injuries.
+It is not only patient level factors that affect how people might be affected<!-- affect and affect--> by trauma.
+About 90% of all trauma related deaths occur in low- and middle-income countries (LMIC), with death rates by trauma also being higher than in high income countries.
+Even within these LMIC, people of poorer socio-economic status have higher death rates from trauma.
 Problems identified within these countries that contribute to these statisitics were infrastructure, education and training, attributed to lack of funding, brain drain to HIC and lack of availability of basic amenities [@Shanthakumar_Payne_Leitch_Alfa-Wali_2021].
@@
 Trauma systems are infrastructures that exist to provide and optimize care for injured patients starting with injury recognition and triage, transport to appropriate trauma center, inpatient care and outpatient follow-up.
-Beyond the clinical side, it works with outreach, education and advocacy, data collection through registries, research, funding, and disaster preparedness and response.
+Beyond the clinical side, it<!-- DO you mean trauma systems?--> works with outreach, education and advocacy, data collection through registries, research, funding, and disaster preparedness and response.
 A comprehensive and functioning trauma system requires strong leadership and engagement at the trauma center, regional and national level [@connolly_woo_lampron_perry_2018].
 This system is crucial to provide care for trauma patients, both in reducing morbidity and mortality in this patient group.
-Earlier studies from Sweden have shown that treating severely injured patients at one of these establishments is associated with a 41% lower adjusted 30-day mortality rate compared to being treated at a non trauma care center due to them being more capable of treating these patients, with patients with higher ISS scores having an even higher protective effect with \>90% for the most critically injured patients [@Candefjord_Asker_Caragounis_2020].
+Earlier studies from Sweden have shown that treating severely injured patients at one of these establishments<!-- Trauma center?--> is associated with a 41% lower adjusted 30-day mortality rate compared to being treated at a non trauma care center due to them being more capable of treating these patients, with patients with higher ISS scores having an even higher protective effect with \>90% <!--90% what?-->for the most critically injured patients [@Candefjord_Asker_Caragounis_2020].
 Other studies have shown similar results, showing that treating injured patients at trauma centers is associated with a 15-30% decrease in mortality [@Moran_Lecky_Bouamra_Lawrence_Edwards_Woodford_Willett_Coats_2018][@Celso_Tepas_Langland-Orban_Pracht_Papa_Lottenberg_Flint_2006][@MacKenzie_Rivara_Jurkovich_Nathens_Frey_Egleston_Salkever_Scharfstein_2006][@Haas_Stukel_Gomez_Zagorski_De_Mestral_Sharma_Rubenfeld_Nathens_2012].
@@
 Trauma teams are multidisciplinary and operate in these centers.
-They play a pivotal role in the initial care of the trauma patient, as they are primary care givers in the critical stage of trauma.
-In Sweden, they consist of individuals from the specialities of surgery, intensive care, orthopedics, nursing and support staff and are lead by a team leader.
-Their aim is to rapidly assess and stabilise the patient, prioritise their injuries and arrange for site of definitive care.
-Emergency nurses are often the personnel activating the trauma code using anatomical and physiological criteria, mechanism of injury and information obtained from emergency service personnel.
-Once in the trauma room, the team works systematically to assess and manage critically ill patients.
-They handle the most urgent problems first, such as airways and breathing, and then tend to less life threatening injuries before readying the patient for transport to the next site of care [@Hedberg_Falkén_Wihlke_Jansson_Hultgren_Eriksson_Beckman_Haupt_Kjörholt_Sandersjöö_et_al._2020].
+They play a pivotal role in the initial care of the trauma patient, as they are primary care givers in the critical stage of trauma<!-- primary care givers feels a bit odd in this context.
+PRovide the initial care perhaps?-->.
+In Sweden, they<!-- Trauma teams?--> include individuals<!-- practitioners?--> from the specialities of surgery, intensive care, orthopedics, nursing and support staff and are lead by a team leader<!-- who is a surgeon-->.
+Their<!-- The trauma team's?--> aim is to rapidly assess and stabilise the patient, prioritise their injuries and arrange for site of definitive care.
+Emergency nurses are often the personnel activating the trauma code<!-- Introduce the concept of a trauma code first--> using anatomical and physiological criteria, mechanism of injury and information obtained from emergency service personnel.
+Once in the trauma room, the team works systematically to assess and manage critically ill patients<!-- Repetition-->.
+They handle the most urgent problems first, such as airways and breathing, and then tend to less life threatening injuries before readying the patient for transport to the next site of care<!-- What do you mean by "next site of care"?--> [@Hedberg_Falkén_Wihlke_Jansson_Hultgren_Eriksson_Beckman_Haupt_Kjörholt_Sandersjöö_et_al._2020].
 
@@
 
-Many hospitals, mainly in HIC, have a registry database where they log the patients information and timeline of what happened.
-This is important work, as a cornerstone of trauma quality improvement programs is multidisciplinary mortality and morbidity review to identify OFIs and implement corrective actions.
+Many hospitals, mainly in HIC, have a registry database<!-- Do you mean trauma registry?--> where they log the patients information and timeline of what happened.
+This is important work, as a cornerstone of trauma quality improvement programs is multidisciplinary mortality and morbidity review to identify OFIs<!-- You have not mentioned OFI before this.
+Introduce the concept of OFI above--> and implement corrective actions.
 This refers to a meeting where different specialities who work with trauma care sit down and go through cases where possible OFIs were identified by nurses who go through the trauma care quality database.
-These MoM conferences are very resource and time craving, which is why many hospitals doesn't have the means to perform them.
-In Sweden
+These MoM conferences are very resource and time craving, which is why many hospitals doesn't<!-- Do not use contractions--> have the means to perform them.
+In Sweden<!-- Something is missing here-->
+
+<!-- I suggest that you introduce the concept of trauma qulity improvement programmes, including mortality and morbidity review and OFI, before you introduce delay in treatment-->
 
@@
 One common OFI that this paper will look further into is delayed treatment.
-Among the preventable or possibly preventable deaths in trauma patients, a subgroup of patients determined by the Combined Trauma Death Review Committee, delay in treatment has been identified as a major error contributing to death, found in up to 52.9% of patients in said group [@Teixeira_Inaba_Hadjizacharia_Brown_Salim_Rhee_Browder_Noguchi_Demetriades_2007].
+Among the preventable or possibly preventable deaths in trauma patients, a subgroup of patients determined by the Combined Trauma Death Review Committee<!-- What's this?
+Elaborate-->, delay in treatment has been identified as a major error contributing to death, found in up to 52.9% of patients in said group [@Teixeira_Inaba_Hadjizacharia_Brown_Salim_Rhee_Browder_Noguchi_Demetriades_2007].
 Although it is such a common OFI, the patient level factors associated with delayed treatment are poorly understood.
 There may be several factors that correlate with receiving delayed treatment, but they have yet to be identified.
-Mapping these factors may help in identifying patients that might be at risk for receiving delayed treatment before it happens.
+Mapping these factors may help in identifying patients that might be at risk for receiving delayed treatment before it happens.<!-- Elaborate on the potential risks of delayed treatment and give examples-->
 
@@
 This is a registry based cohort study using data from the trauma registry and trauma care quality database at the Karolinska University Hospital in Solna.
-The trauma registry includes about 12000 patients treated between 2012 and 2022.
-The trauma care quality database is a subset of the trauma registry and includes about 6000 patients selected for review between 2013 and 2022.
-The outcome is delayed treatment, as identified by the multidisciplinary review board and recorded in the trauma care quality database.
+The trauma registry includes about 14000 patients treated between 2012 and 2022.
+The trauma care quality database is a subset of the trauma registry and includes about 8000 patients selected for review between 2013 and 2022.
+The outcome is delayed treatment<!-- You can have a separate heading called Outcome-->, as identified by the multidisciplinary review board and recorded in the trauma care quality database.
 This project will link the two databases and assess how different patient level factors, such as age, sex, mechanism of injury, and injury severity, are associated with delayed treatment using logistic regression.
-A 5% significance level and 95% confidence levels will be used.
+A 5% significance level and 95% confidence levels will be used.<!-- This last line should be under statistical methods-->
 
@@
 
-Every year, Karolinska University hospital, which is the equivalent of a trauma level 1 hospital, admits 1500 trauma patients each year to their facility.
+Every year, Karolinska University hospital, which is the equivalent of a trauma level 1 hospital, admits 1500 trauma patients each year to their facility<!-- Last part is repetition-->.
 To be added to the trauma registry, a patient must be over 15 years of age, had ISS\>9 and/or had trauma code activation.
 They get added to both the Karolinska Trauma registry but also the national trauma registry (SweTrau).
-The registry includes data on vital signs, injuries, interventions and patient information.
-A subset to this database, he trauma care quality database includes information relevant for mortality and morbidity conferences
+The trauma registry includes data on vital signs, injuries, interventions and patient information.
+A subset to this database, the trauma care quality database<!-- Use more direct language: The trauma care quality database is a subset...--> includes information relevant for mortality and morbidity conferences
 
@@
 Inclusion in the study further requires that the patients have been assessed for OFIs.
-Exclusion happens if a patient is missing information in one of the examined patient factors.
+Exclusion happens<!-- Same here, use direct language: Patients were excluded if...--> if a patient is missing information in one of the examined patient factors.
 
@@
 This project will link the two databases and assess how different patient level factors, such as age, sex, mechanism of injury, and injury severity, are associated with delayed treatment using logistic regression.
-A 5% significance level and 95% confidence levels will be used.
+A 5% significance level and 95% confidence levels will be used.<!-- Did you forget to remove these two last sentence?-->
 
@@
 
-Bias is not relevant in this study as the data is collected from a database and processed by a coding program.
+Bias is not relevant in this study as the data is collected from a database and processed by a coding program.<!-- Not quite true, there are multiple biases that may come into play, for example selection bias, i.e.
+is the sample representative of the population to which you want to generalise?-->
 

### 2023-09-20 manuscriptlongversion.Rmd

@@
 These MoM conferences are very resource and time craving, which is why many hospitals doesn't<!-- Do not use contractions--> have the means to perform them.
-In Sweden<!-- Something is missing here-->
+In Sweden In Sweden, most hospitals have mortality conferences on deceased patient cases, but only one has MoM conferences to better their trauma care on patients who lived.
 
@@
 
-Every year, Karolinska University hospital, which is the equivalent of a trauma level 1 hospital, admits 1500 trauma patients each year to their facility<!-- Last part is repetition-->.
+Every year, Karolinska University hospital, which is the equivalent of a trauma level 1 hospital, admits 1500 trauma patients each year to their facility.
 To be added to the trauma registry, a patient must be over 15 years of age, had ISS\>9 and/or had trauma code activation.
 They get added to both the Karolinska Trauma registry but also the national trauma registry (SweTrau).
-The trauma registry includes data on vital signs, injuries, interventions and patient information.
-A subset to this database, the trauma care quality database<!-- Use more direct language: The trauma care quality database is a subset...--> includes information relevant for mortality and morbidity conferences
+The registry includes data on vital signs, injuries, interventions and patient information.
+A subset to this database, the trauma care quality database includes information relevant for MoM conferences.
+At Karolinska, patients who die during hospital stay gets their case sent to a MoM everytime.
+However, for the rest of the patients, it is determined by two specialized nurses.
+Several audit filters are set up to flag for patients with possible OFIs.
+Flagged patients are then reviewed by these two nurses who decide if there is a possible OFI, in which case their case is sent to a MoM conference for further review.
 

### 2023-09-21 manuscriptlongversion.Rmd
intro

@@
 
-Within modern hospital care, trauma care stands as an important part in healthcare and is crucial to lowering mortality and morbidity in injured patients<!-- Not only hospital trauma care, you may want to mention the entire system already here-->.
-The definition of trauma is "a set of psychological and physiological responses to pain, injury, serious illness, medical procedures and frightening treatment experiences" [@International_Society_for_Traumatic_Stress_Studies_2020].
-As the definition suggests, it encompasses a broad spectrum of different conditions, both physical and psychological<!-- I don't think you need this definition here, it doesn't really fit-->.
-The one we will be focusing on in this paper is physical trauma, the clinical entity composed of physical injury and the body's associated response, hereby referred to as trauma.
+Trauma, defined as the clinical entity composed of physical injury and the body's associated response, is and has long been a major cause of death around the world today.
+
+Within modern hospital care, trauma systems stands as an important part in healthcare and is crucial to lowering mortality and morbidity in injured patients through many means, such as hospital care, follow up and prevention programs.
 
@@
 
-It is often divided into two subgroups<!-- By "it", do you mean trauma?
-Suggest you just write so-->, blunt force trauma and penetrating trauma.
+Trauma is often divided into two subgroups, blunt force trauma and penetrating trauma.
 Blunt force trauma is when an object or force strikes the body, often causing bruising, broken bones or deep cuts.
-Examples of this could be car crashes, falls or direct blows to the body.
-Penetrating<!-- More common to refer to it as penetrating, rather than penetrative--> trauma is when an object pierces the skin or body and creates an open wound.
-Examples of this<!-- By "this", do you mean penetrating trauma mechanisms?--> are gunshot wounds and stab wounds [@National_Institute_of_General_Medical_Sciences_2020].
-Differences exist between the two groups, as blunt trauma patients tend to score higher on ISS scale<!-- You haven't mentioned ISS before here.
-Do you mean that blunt trauma patients are more severely injured?--> on arrival and as such require more resources and be hospitalized for a longer period because of their more severe injuries.
+Examples of blunt trauma could be car crashes, falls or direct blows to the body.
+Penetrating trauma is when an object pierces the skin or body and creates an open wound.
+Examples of penetrating trauma are gunshot wounds and stab wounds [@National_Institute_of_General_Medical_Sciences_2020].
+Differences exist between the two groups, as blunt trauma patients tend to be more injured on arrival to the hospital and as such require more resources and be hospitalized for a longer period because of their more severe injuries [@FITCH20191201].
 
-Trauma is and has long been a major cause of death around the world today, taking the lives of around 4,4 million people each year, almost 8% of all deaths [@World_Health_Organization_2021].
+Today, trauma takes the lives of around 4,4 million people each year, almost 8% of all deaths [@World_Health_Organization_2021].
 In the United States, trauma is the 4th leading cause of death among the general population and the leading cause for people between the ages of 1 and 44 [@AAST_2020].
@@
 This has a significant burden on social and economic level, costing countries billions of US dollars each year in healthcare and lost productivity[@connolly_woo_lampron_perry_2018].
-Studies estimate the cost of trauma care being between 18,500\$ and 41,500\$ in HIC<!-- Spell out the first time you mention HIC-->, depending on the country [@Willenberg_Curtis_Taylor_Jan_Glass_Myburgh_2012].
+Studies estimate the cost of trauma care being between 18,500\$ and 41,500\$ in high income countries, depending on the country [@Willenberg_Curtis_Taylor_Jan_Glass_Myburgh_2012].
 However, as one might expect<!-- Phrases like "as one might expect", "it is known that", "as has been shown" are usually unecessary and don't add much.

### 2023-09-21 manuscriptlongversion.Rmd
intro

@@
 Studies estimate the cost of trauma care being between 18,500\$ and 41,500\$ in high income countries, depending on the country [@Willenberg_Curtis_Taylor_Jan_Glass_Myburgh_2012].
-However, as one might expect<!-- Phrases like "as one might expect", "it is known that", "as has been shown" are usually unecessary and don't add much.
-Only use them if you need the extra words -->, this burden is not evenly distributed between or within countries.
+However, it has been shown that this burden is not evenly distributed between or within countries.
 Many social factors such as age, sex and social status play a major role in the risk of dying from trauma, with young men with low socio-economic status being at most risk [@World_Health_Organization_2021].
-It is not only patient level factors that affect how people might be affected<!-- affect and affect--> by trauma.
+But it is not only patient level factors that affect what effect trauma has on people.
 About 90% of all trauma related deaths occur in low- and middle-income countries (LMIC), with death rates by trauma also being higher than in high income countries.
@@
 Trauma systems are infrastructures that exist to provide and optimize care for injured patients starting with injury recognition and triage, transport to appropriate trauma center, inpatient care and outpatient follow-up.
-Beyond the clinical side, it<!-- DO you mean trauma systems?--> works with outreach, education and advocacy, data collection through registries, research, funding, and disaster preparedness and response.
+Beyond the clinical side, trauma systems work with outreach, education and advocacy, data collection through registries, research, funding, and disaster preparedness and response.
 A comprehensive and functioning trauma system requires strong leadership and engagement at the trauma center, regional and national level [@connolly_woo_lampron_perry_2018].
 This system is crucial to provide care for trauma patients, both in reducing morbidity and mortality in this patient group.
-Earlier studies from Sweden have shown that treating severely injured patients at one of these establishments<!-- Trauma center?--> is associated with a 41% lower adjusted 30-day mortality rate compared to being treated at a non trauma care center due to them being more capable of treating these patients, with patients with higher ISS scores having an even higher protective effect with \>90% <!--90% what?-->for the most critically injured patients [@Candefjord_Asker_Caragounis_2020].
+Earlier studies from Sweden have shown that treating severely injured patients at a trauma center is associated with a 41% lower adjusted 30-day mortality rate compared to being treated at a non trauma care center due to them being more capable of treating these patients, with potential survival benefit increasing with higher injury severity [@Candefjord_Asker_Caragounis_2020].
 Other studies have shown similar results, showing that treating injured patients at trauma centers is associated with a 15-30% decrease in mortality [@Moran_Lecky_Bouamra_Lawrence_Edwards_Woodford_Willett_Coats_2018][@Celso_Tepas_Langland-Orban_Pracht_Papa_Lottenberg_Flint_2006][@MacKenzie_Rivara_Jurkovich_Nathens_Frey_Egleston_Salkever_Scharfstein_2006][@Haas_Stukel_Gomez_Zagorski_De_Mestral_Sharma_Rubenfeld_Nathens_2012].
@@
 Trauma teams are multidisciplinary and operate in these centers.
-They play a pivotal role in the initial care of the trauma patient, as they are primary care givers in the critical stage of trauma<!-- primary care givers feels a bit odd in this context.
-PRovide the initial care perhaps?-->.
-In Sweden, they<!-- Trauma teams?--> include individuals<!-- practitioners?--> from the specialities of surgery, intensive care, orthopedics, nursing and support staff and are lead by a team leader<!-- who is a surgeon-->.
-Their<!-- The trauma team's?--> aim is to rapidly assess and stabilise the patient, prioritise their injuries and arrange for site of definitive care.
-Emergency nurses are often the personnel activating the trauma code<!-- Introduce the concept of a trauma code first--> using anatomical and physiological criteria, mechanism of injury and information obtained from emergency service personnel.
-Once in the trauma room, the team works systematically to assess and manage critically ill patients<!-- Repetition-->.
-They handle the most urgent problems first, such as airways and breathing, and then tend to less life threatening injuries before readying the patient for transport to the next site of care<!-- What do you mean by "next site of care"?--> [@Hedberg_Falkén_Wihlke_Jansson_Hultgren_Eriksson_Beckman_Haupt_Kjörholt_Sandersjöö_et_al._2020].
+They play a pivotal role in the treatment of the trauma patient, as they provide the initial care in the critical stage of trauma.
+In Sweden, trauma teams are lead by a team leader who is a surgeon, and include practitioners from the specialities of intensive care, orthopedics, nursing and support staff.
+For the trauma team to be mobilized, trauma code has to be activated, often by an emergency nurse.
+The nurse uses information gathered by the first responders to see if the patient fits any of the criteria for trauma code.
+Most healthcare facilities have established criteria or guidelines that trigger trauma code activation.
+These criteria typically include mechanisms of injury, physiological criteria, anatomical criteria and other specific indicators of severe trauma.
+There are different levels of trauma code with different criteria correlating with severity of injury, with level 1 mobilizing the most personnel to the trauma room.
+Once in the trauma room, the team works systematically to manage the patients injuries.
+They handle the most urgent problems first, such as airways and breathing, with the aim to rapidly assess and stabilise the patient, prioritise their injuries and arrange for site of definitive care.
+[@Hedberg_Falkén_Wihlke_Jansson_Hultgren_Eriksson_Beckman_Haupt_Kjörholt_Sandersjöö_et_al._2020].
 
-**Delay in treatment**
+**Mortality and morbidity conferences**
 
-Many hospitals, mainly in HIC, have a registry database<!-- Do you mean trauma registry?--> where they log the patients information and timeline of what happened.
-This is important work, as a cornerstone of trauma quality improvement programs is multidisciplinary mortality and morbidity review to identify OFIs<!-- You have not mentioned OFI before this.
-Introduce the concept of OFI above--> and implement corrective actions.
+Many hospitals, mainly in HIC, have a trauma registry where they log the patients information and timeline of what happened.
+This is important work, as a cornerstone of trauma quality improvement programs is multidisciplinary mortality and morbidity review to identify opportunities for improvement and implement corrective actions.
+OFIs are the endpoint of MoM conferences.
+Once
+
 This refers to a meeting where different specialities who work with trauma care sit down and go through cases where possible OFIs were identified by nurses who go through the trauma care quality database.
@@
 In Sweden In Sweden, most hospitals have mortality conferences on deceased patient cases, but only one has MoM conferences to better their trauma care on patients who lived.
+Examples of opportunities for improvement may include lack of resources and management errors.
 
@@
 
-Examples of opportunities for improvement may include lack of resources and management errors.
 One common OFI that this paper will look further into is delayed treatment.
@@
 This project will link the two databases and assess how different patient level factors, such as age, sex, mechanism of injury, and injury severity, are associated with delayed treatment using logistic regression.
-A 5% significance level and 95% confidence levels will be used.<!-- This last line should be under statistical methods-->
 
@@
 Inclusion in the study further requires that the patients have been assessed for OFIs.
-Exclusion happens<!-- Same here, use direct language: Patients were excluded if...--> if a patient is missing information in one of the examined patient factors.
+Patients were excluded if they were missing information in one of the examined patient factors.
 
@@
 The outcome is delayed treatment, which is defined as more than 30 minutes passing from arrival before receiving treatment, as identified by the multidisciplinary review board and recorded in the trauma care quality database.
-This project will link the two databases and assess how different patient level factors, such as age, sex, mechanism of injury, and injury severity, are associated with delayed treatment using logistic regression.
-A 5% significance level and 95% confidence levels will be used.<!-- Did you forget to remove these two last sentence?-->
 
@@
 
-Multivariable and potentially univariable logistical regression
+Multivariable and potentially univariable logistical regression. A 5% significance level and 95% confidence levels will be used.

### 2023-09-21 manuscriptlongversion.Rmd
intro

@@
 This has a significant burden on social and economic level, costing countries billions of US dollars each year in healthcare and lost productivity[@connolly_woo_lampron_perry_2018].
-Studies estimate the cost of trauma care being between 18,500\$ and 41,500\$ in high income countries, depending on the country [@Willenberg_Curtis_Taylor_Jan_Glass_Myburgh_2012].
+Studies estimate the cost of trauma care being between 18,500\$ and 41,500\$ in high income countries (HIC), depending on the country [@Willenberg_Curtis_Taylor_Jan_Glass_Myburgh_2012].
 However, it has been shown that this burden is not evenly distributed between or within countries.
@@
 
-**Mortality and morbidity conferences**
+**Mortality and morbidity (MoM) conferences**
 
 Many hospitals, mainly in HIC, have a trauma registry where they log the patients information and timeline of what happened.
-This is important work, as a cornerstone of trauma quality improvement programs is multidisciplinary mortality and morbidity review to identify opportunities for improvement and implement corrective actions.
-OFIs are the endpoint of MoM conferences.
-Once
+This is important work, as a cornerstone of trauma quality improvement programs is multidisciplinary MoM conferences.
+The MoM conference is a meeting where different specialities who work with trauma care sit down and discuss deaths and complications in order to look for preventable factors.
+These conferences are performed in many hospitals globally, almost everywhere where there are formal medical specialty departments and sometimes in smaller hospitals as well [@World_Health_Organization_2009].
 
-This refers to a meeting where different specialities who work with trauma care sit down and go through cases where possible OFIs were identified by nurses who go through the trauma care quality database.
-These MoM conferences are very resource and time craving, which is why many hospitals doesn't<!-- Do not use contractions--> have the means to perform them.
+OFIs are the endpoint of MoM conferences. These meetings are
+
+where possible opportunities for improvement (OFI) were identified by nurses who go through the trauma care quality database
+
+These MoM conferences are very resource and time craving, which is why it is a concept that has not been broadly implemented.
+
 In Sweden In Sweden, most hospitals have mortality conferences on deceased patient cases, but only one has MoM conferences to better their trauma care on patients who lived.

### 2023-09-21 manuscriptlongversion.Rmd
intro

@@
 
-Many hospitals, mainly in HIC, have a trauma registry where they log the patients information and timeline of what happened.
-This is important work, as a cornerstone of trauma quality improvement programs is multidisciplinary MoM conferences.
+Many hospitals have a trauma registry where they log the patients information and timeline of what happened.
+The extent of the registry varies, with HIC having more complete registries.
+This logging of information is important work, as a cornerstone of trauma quality improvement programs is multidisciplinary MoM conferences.
 The MoM conference is a meeting where different specialities who work with trauma care sit down and discuss deaths and complications in order to look for preventable factors.
+The endpoint of these MoM conferences are opportunities for improvement (OFI).
+At the end of the conference, consensus is reached regarding the existence of an OFI.
+
 These conferences are performed in many hospitals globally, almost everywhere where there are formal medical specialty departments and sometimes in smaller hospitals as well [@World_Health_Organization_2009].
 
-OFIs are the endpoint of MoM conferences. These meetings are
+In Sweden, most hospitals have mortality conferences on deceased patient cases, but only one has MoM conferences to better their trauma care on patients who lived.
+The reason being that this expansion of MoM conferences takes time and resources, which is why all hospitals don't implement this.
 
-where possible opportunities for improvement (OFI) were identified by nurses who go through the trauma care quality database
-
-These MoM conferences are very resource and time craving, which is why it is a concept that has not been broadly implemented.
-
-In Sweden In Sweden, most hospitals have mortality conferences on deceased patient cases, but only one has MoM conferences to better their trauma care on patients who lived.
-Examples of opportunities for improvement may include lack of resources and management errors.
-
-<!-- I suggest that you introduce the concept of trauma qulity improvement programmes, including mortality and morbidity review and OFI, before you introduce delay in treatment-->
+Examples of OFI may include lack of resources and management errors.
 

### 2023-09-21 manuscriptlongversion.Rmd
intro

@@
 The MoM conference is a meeting where different specialities who work with trauma care sit down and discuss deaths and complications in order to look for preventable factors.
-The endpoint of these MoM conferences are opportunities for improvement (OFI).
-At the end of the conference, consensus is reached regarding the existence of an OFI.
-
 These conferences are performed in many hospitals globally, almost everywhere where there are formal medical specialty departments and sometimes in smaller hospitals as well [@World_Health_Organization_2009].
-
 In Sweden, most hospitals have mortality conferences on deceased patient cases, but only one has MoM conferences to better their trauma care on patients who lived.
@@
 
+The endpoint of these MoM conferences are opportunities for improvement (OFI).
+At the end of the conference, consensus is reached regarding the existence of an OFI and implementation of corrective action.
+This process is effective, as it has been shown that this review is associated with high-quality trauma care [@Santana_Stelfox_2014].
 Examples of OFI may include lack of resources and management errors.
-
 One common OFI that this paper will look further into is delayed treatment.
-Among the preventable or possibly preventable deaths in trauma patients, a subgroup of patients determined by the Combined Trauma Death Review Committee<!-- What's this?
-Elaborate-->, delay in treatment has been identified as a major error contributing to death, found in up to 52.9% of patients in said group [@Teixeira_Inaba_Hadjizacharia_Brown_Salim_Rhee_Browder_Noguchi_Demetriades_2007].
+According to previous studies, among the preventable or possibly preventable deaths in trauma patients, delay in treatment has been identified as a major error contributing to death, found in up to 52.9% of patients in said group [@Teixeira_Inaba_Hadjizacharia_Brown_Salim_Rhee_Browder_Noguchi_Demetriades_2007].
 Although it is such a common OFI, the patient level factors associated with delayed treatment are poorly understood.

### 2023-09-22 manuscriptlongversion.Rmd
intro
intro and method

@@
 Trauma, defined as the clinical entity composed of physical injury and the body's associated response, is and has long been a major cause of death around the world today.
-
-Within modern hospital care, trauma systems stands as an important part in healthcare and is crucial to lowering mortality and morbidity in injured patients through many means, such as hospital care, follow up and prevention programs.
+Within modern hospital care, the trauma system stands as an important part in healthcare and is crucial to lowering mortality and morbidity in injured patients through many means, such as hospital care, follow up and prevention programs.
 
@@
 These conferences are performed in many hospitals globally, almost everywhere where there are formal medical specialty departments and sometimes in smaller hospitals as well [@World_Health_Organization_2009].
-In Sweden, most hospitals have mortality conferences on deceased patient cases, but only one has MoM conferences to better their trauma care on patients who lived.
+In Sweden, many hospitals have mortality conferences on deceased patient cases, but only one has MoM conferences to better their trauma care on patients who lived [@SweTrau_2023].
 The reason being that this expansion of MoM conferences takes time and resources, which is why all hospitals don't implement this.
@@
 According to previous studies, among the preventable or possibly preventable deaths in trauma patients, delay in treatment has been identified as a major error contributing to death, found in up to 52.9% of patients in said group [@Teixeira_Inaba_Hadjizacharia_Brown_Salim_Rhee_Browder_Noguchi_Demetriades_2007].
+Delayed treatment has been shown to have adverse effects on patients, showing why it is such an important issue and the need to develop strategies to combat [@Sampalis_Boukas_Lavoie_Nikolis_Frechette_Brown_Fleiszer_Mulder_1995][@Sampalis_Lavoie_Salas_Nikolis_Williams_1994].
 Although it is such a common OFI, the patient level factors associated with delayed treatment are poorly understood.
 There may be several factors that correlate with receiving delayed treatment, but they have yet to be identified.
-Mapping these factors may help in identifying patients that might be at risk for receiving delayed treatment before it happens.<!-- Elaborate on the potential risks of delayed treatment and give examples-->
+Mapping these factors may help in identifying patients that might be at risk for receiving delayed treatment before it happens.
+
+**Aim**
 
@@
 The trauma care quality database is a subset of the trauma registry and includes about 8000 patients selected for review between 2013 and 2022.
-The outcome is delayed treatment<!-- You can have a separate heading called Outcome-->, as identified by the multidisciplinary review board and recorded in the trauma care quality database.
 This project will link the two databases and assess how different patient level factors, such as age, sex, mechanism of injury, and injury severity, are associated with delayed treatment using logistic regression.
+
+**Outcome**
+
+The outcome is delayed treatment, as identified by the multidisciplinary review board and recorded in the trauma care quality database.
 
@@
 
-Bias is not relevant in this study as the data is collected from a database and processed by a coding program.<!-- Not quite true, there are multiple biases that may come into play, for example selection bias, i.e.
-is the sample representative of the population to which you want to generalise?-->
+Selection bias as this is a retrospective cohort study where the outcome has occurred.
+
+Misclassification bias as the conference may reach a consensus that is wrong.
 
@@
 
-**Quantitative variables**
-
 **Statistical methods**

### 2023-09-23 manuscriptlongversion.Rmd

@@
 bibliography: bibliography.bib
+editor_options:
+  markdown:
+    wrap: 72
 ---
@@
 
-Trauma, defined as the clinical entity composed of physical injury and the body's associated response, is and has long been a major cause of death around the world today.
-Within modern hospital care, the trauma system stands as an important part in healthcare and is crucial to lowering mortality and morbidity in injured patients through many means, such as hospital care, follow up and prevention programs.
+Trauma, defined as the clinical entity composed of physical injury and
+the body's associated response <!-- KS: A good citation is:
+@book{gerdin_risk_2015,
+    location = {Stockholm},
+    title = {The risk of dying: predicting trauma mortality in urban Indian
+             hospitals},
+    isbn = {978-91-7676-157-1},
+    shorttitle = {The risk of dying},
+    publisher = {Karolinska Institutet},
+    author = {Gerdin, Martin},
+    date = {2015},
+    note = {{OCLC}: 942942790},
+}
+A good read if you have time ;), PDF URL: https://openarchive.ki.se/xmlui/bitstream/handle/10616/44832/Thesis_Martin_Gerdin.pdf?sequence=3&isAllowed=y
+--> , is and has long been a major cause of death around the world today
+<!-- KS: needs citation -->. Within modern hospital care, the trauma
+system stands as an important part in healthcare and is crucial to
+lowering mortality and morbidity <!-- KS: needs citation --> in injured
+patients through many means, such as hospital care, follow up and
+prevention programs.
 
-**Trauma statistics**
+<!-- KS: Use markdown headings. So here you should do the below: -->
 
-Trauma is often divided into two subgroups, blunt force trauma and penetrating trauma.
-Blunt force trauma is when an object or force strikes the body, often causing bruising, broken bones or deep cuts.
-Examples of blunt trauma could be car crashes, falls or direct blows to the body.
-Penetrating trauma is when an object pierces the skin or body and creates an open wound.
-Examples of penetrating trauma are gunshot wounds and stab wounds [@National_Institute_of_General_Medical_Sciences_2020].
-Differences exist between the two groups, as blunt trauma patients tend to be more injured on arrival to the hospital and as such require more resources and be hospitalized for a longer period because of their more severe injuries [@FITCH20191201].
+## Trauma Statistics
 
-Today, trauma takes the lives of around 4,4 million people each year, almost 8% of all deaths [@World_Health_Organization_2021].
-In the United States, trauma is the 4th leading cause of death among the general population and the leading cause for people between the ages of 1 and 44 [@AAST_2020].
-It is also an important cause of hospitalization and morbidity among all age groups, including seniors, and is responsible for an estimated 10% of all years lived with a disability globally.
-This has a significant burden on social and economic level, costing countries billions of US dollars each year in healthcare and lost productivity[@connolly_woo_lampron_perry_2018].
-Studies estimate the cost of trauma care being between 18,500\$ and 41,500\$ in high income countries (HIC), depending on the country [@Willenberg_Curtis_Taylor_Jan_Glass_Myburgh_2012].
-However, it has been shown that this burden is not evenly distributed between or within countries.
-Many social factors such as age, sex and social status play a major role in the risk of dying from trauma, with young men with low socio-economic status being at most risk [@World_Health_Organization_2021].
-But it is not only patient level factors that affect what effect trauma has on people.
-About 90% of all trauma related deaths occur in low- and middle-income countries (LMIC), with death rates by trauma also being higher than in high income countries.
-Even within these LMIC, people of poorer socio-economic status have higher death rates from trauma.
-Problems identified within these countries that contribute to these statisitics were infrastructure, education and training, attributed to lack of funding, brain drain to HIC and lack of availability of basic amenities [@Shanthakumar_Payne_Leitch_Alfa-Wali_2021].
-In HIC where funding and governance over the healthcare system is better designed, functioning trauma systems and dedicated trauma centers exist that have been shown to lower mortality but also improve functional outcome in trauma patients [@Nirula_Brasel_2006].
+and not: **Trauma statistics** <!-- KS: use title casing -->
+<!-- KS: personally if feel you are writing too much background to trauma.
+You should focus on what your aim is.
+Delay in trauma treatment.
+Thus write very little about trauma in general and more about trauma treatment and quality improvement programmes.
+You are not writing a review article but a specific study. Thus information such as cost etc can be ignored. One paragraph on statistic is fine -->
+
+Trauma is often divided into two subgroups, blunt force trauma and
+penetrating trauma. Blunt force trauma is when an object or force
+strikes the body, often causing bruising, broken bones or deep cuts.
+Examples of blunt trauma could be car crashes, falls or direct blows to
+the body. Penetrating trauma is when an object pierces the skin or body
+and creates an open wound. Examples of penetrating trauma are gunshot
+wounds and stab wounds
+[@National_Institute_of_General_Medical_Sciences_2020]
+<!-- KS: Please use shorter bib keys. Author and year is enough, ex. [@szolnoky2023] -->.
+Differences exist between the two groups, as blunt trauma patients tends
+to be more injured on arrival to the hospital and as such require more
+resources and be hospitalized for a longer period because of their more
+severe injuries [@FITCH20191201] <!-- KS: reword, does not flow -->.
+
+Today, trauma takes the lives of around 4,4 million people each year,
+almost 8% of all deaths [@World_Health_Organization_2021]. In the United
+States, trauma is the 4th leading cause of death among the general
+population and the leading cause for people between the ages of 1 and 44
+[@AAST_2020]. It is also an important cause of hospitalization and
+morbidity among all age groups, including seniors, and is responsible
+for an estimated 10% of all years lived with a disability globally. This
+has a significant burden on social and economic level, costing countries
+billions of US dollars each year in healthcare and lost
+productivity[@connolly_woo_lampron_perry_2018]. Studies estimate the
+cost of trauma care being between 18,500\$ and 41,500\$ in high income
+countries (HIC), depending on the country
+<!-- KS: per patient or per what? -->
+[@Willenberg_Curtis_Taylor_Jan_Glass_Myburgh_2012]. However, it has been
+shown that this burden is not evenly distributed between or within
+countries. Many social factors such as age, sex and social status play a
+major role in the risk of dying from trauma, with young men with low
+socio-economic status being at most risk
+[@World_Health_Organization_2021]. But it is not only patient level
+factors that affect what effect trauma has on people. About 90% of all
+trauma related deaths occur in low- and middle-income countries (LMIC),
+with death rates by trauma also being higher than in high income
+countries. Even within these LMIC, people of poorer socio-economic
+status have higher death rates from trauma. Problems identified within
+these countries that contribute to these statisitics were
+infrastructure, education and training, attributed to lack of funding,
+brain drain to HIC and lack of availability of basic amenities
+[@Shanthakumar_Payne_Leitch_Alfa-Wali_2021]. In HIC where funding and
+governance over the healthcare system is better designed, functioning
+trauma systems and dedicated trauma centers exist that have been shown
+to lower mortality but also improve functional outcome in trauma
+patients [@Nirula_Brasel_2006].
 
 **Trauma systems**
+<!-- KS: Keep in mind that your target audience is your peers. They do have some
+medical knowledge. Thus you do not have to define everything and can shorten 
+this section. -->
 
-Trauma systems are infrastructures that exist to provide and optimize care for injured patients starting with injury recognition and triage, transport to appropriate trauma center, inpatient care and outpatient follow-up.
-Beyond the clinical side, trauma systems work with outreach, education and advocacy, data collection through registries, research, funding, and disaster preparedness and response.
-A comprehensive and functioning trauma system requires strong leadership and engagement at the trauma center, regional and national level [@connolly_woo_lampron_perry_2018].
-This system is crucial to provide care for trauma patients, both in reducing morbidity and mortality in this patient group.
-Earlier studies from Sweden have shown that treating severely injured patients at a trauma center is associated with a 41% lower adjusted 30-day mortality rate compared to being treated at a non trauma care center due to them being more capable of treating these patients, with potential survival benefit increasing with higher injury severity [@Candefjord_Asker_Caragounis_2020].
-Other studies have shown similar results, showing that treating injured patients at trauma centers is associated with a 15-30% decrease in mortality [@Moran_Lecky_Bouamra_Lawrence_Edwards_Woodford_Willett_Coats_2018][@Celso_Tepas_Langland-Orban_Pracht_Papa_Lottenberg_Flint_2006][@MacKenzie_Rivara_Jurkovich_Nathens_Frey_Egleston_Salkever_Scharfstein_2006][@Haas_Stukel_Gomez_Zagorski_De_Mestral_Sharma_Rubenfeld_Nathens_2012].
-During the last three decades, the introduction of trauma systems has contributed to lowering the incidence of preventable death .
-This is attributed to the improvements in care for acute brain injuries and bleeding control.
-The incidence of late death because of sepsis and multiple organ failure, possibly a result of better and earlier resuscitation [@Asensio_Trunkey_2008].
-Functioning trauma systems saves lives in the hospital, but their work outside of the hospital is just as important.
-Data collection from injured patients, such as mechanism of injury or mortality, are essential for creating databases that can be used for research.
-In turn, that research can be used for planning injury prevention programs that target the most common injuries in the right ways, e.g.
-teen drivers, children, specific occupations etc.
-These injury prevention programs can be planned on trauma center, organizational or government level[@ACS].
+Trauma systems are infrastructures that exist to provide and optimize
+care for injured patients starting with injury recognition and triage,
+transport to appropriate trauma center, inpatient care and outpatient
+follow-up. Beyond the clinical side, trauma systems work with outreach,
+education and advocacy, data collection through registries, research,
+funding, and disaster preparedness and response. A comprehensive and
+functioning trauma system requires strong leadership and engagement at
+the trauma center, regional and national level
+[@connolly_woo_lampron_perry_2018]. This system is crucial to provide
+care for trauma patients, both in reducing morbidity and mortality in
+this patient group. Earlier studies from Sweden have shown that treating
+severely injured patients at a trauma center is associated with a 41%
+lower adjusted 30-day mortality rate compared to being treated at a non
+trauma care center due to them being more capable of treating these
+patients, with potential survival benefit increasing with higher injury
+severity [@Candefjord_Asker_Caragounis_2020]. Other studies have shown
+similar results, showing that treating injured patients at trauma
+centers is associated with a 15-30% decrease in mortality
+[@Moran_Lecky_Bouamra_Lawrence_Edwards_Woodford_Willett_Coats_2018][@Celso_Tepas_Langland-Orban_Pracht_Papa_Lottenberg_Flint_2006][@MacKenzie_Rivara_Jurkovich_Nathens_Frey_Egleston_Salkever_Scharfstein_2006][@Haas_Stukel_Gomez_Zagorski_De_Mestral_Sharma_Rubenfeld_Nathens_2012].
+During the last three decades, the introduction of trauma systems has
+contributed to lowering the incidence of preventable death . This is
+attributed to the improvements in care for acute brain injuries and
+bleeding control. The incidence of late death because of sepsis and
+multiple organ failure, possibly a result of better and earlier
+resuscitation [@Asensio_Trunkey_2008]. Functioning trauma systems saves
+lives in the hospital, but their work outside of the hospital is just as
+important. Data collection from injured patients, such as mechanism of
+injury or mortality, are essential for creating databases that can be
+used for research. In turn, that research can be used for planning
+injury prevention programs that target the most common injuries in the
+right ways, e.g. teen drivers, children, specific occupations etc. These
+injury prevention programs can be planned on trauma center,
+organizational or government level[@ACS].
 
-Trauma teams are multidisciplinary and operate in these centers.
-They play a pivotal role in the treatment of the trauma patient, as they provide the initial care in the critical stage of trauma.
-In Sweden, trauma teams are lead by a team leader who is a surgeon, and include practitioners from the specialities of intensive care, orthopedics, nursing and support staff.
-For the trauma team to be mobilized, trauma code has to be activated, often by an emergency nurse.
-The nurse uses information gathered by the first responders to see if the patient fits any of the criteria for trauma code.
-Most healthcare facilities have established criteria or guidelines that trigger trauma code activation.
-These criteria typically include mechanisms of injury, physiological criteria, anatomical criteria and other specific indicators of severe trauma.
-There are different levels of trauma code with different criteria correlating with severity of injury, with level 1 mobilizing the most personnel to the trauma room.
-Once in the trauma room, the team works systematically to manage the patients injuries.
-They handle the most urgent problems first, such as airways and breathing, with the aim to rapidly assess and stabilise the patient, prioritise their injuries and arrange for site of definitive care.
-[@Hedberg_Falkén_Wihlke_Jansson_Hultgren_Eriksson_Beckman_Haupt_Kjörholt_Sandersjöö_et_al._2020].
+Trauma teams are multidisciplinary and operate in these centers. They
+play a pivotal role in the treatment of the trauma patient, as they
+provide the initial care in the critical stage of trauma. In Sweden,
+trauma teams are lead by a team leader who is a surgeon, and include
+practitioners from the specialities of intensive care, orthopedics,
+nursing and support staff. For the trauma team to be mobilized, trauma
+code has to be activated, often by an emergency nurse. The nurse uses
+information gathered by the first responders to see if the patient fits
+any of the criteria for trauma code. Most healthcare facilities have
+established criteria or guidelines that trigger trauma code activation.
+These criteria typically include mechanisms of injury, physiological
+criteria, anatomical criteria and other specific indicators of severe
+trauma. There are different levels of trauma code with different
+criteria correlating with severity of injury, with level 1 mobilizing
+the most personnel to the trauma room. Once in the trauma room, the team
+works systematically to manage the patients injuries. They handle the
+most urgent problems first, such as airways and breathing, with the aim
+to rapidly assess and stabilise the patient, prioritise their injuries
+and arrange for site of definitive care.
+[@Hedberg_Falkén_Wihlke_Jansson_Hultgren_Eriksson_Beckman_Haupt_Kjörholt_Sandersjöö_et_al._2020]
+<!-- KS: bibtex does not like special characters, thus keep the keys short and only use basic letters a-z and 0-9 -->.
 
@@
 
-Many hospitals have a trauma registry where they log the patients information and timeline of what happened.
-The extent of the registry varies, with HIC having more complete registries.
-This logging of information is important work, as a cornerstone of trauma quality improvement programs is multidisciplinary MoM conferences.
-The MoM conference is a meeting where different specialities who work with trauma care sit down and discuss deaths and complications in order to look for preventable factors.
-These conferences are performed in many hospitals globally, almost everywhere where there are formal medical specialty departments and sometimes in smaller hospitals as well [@World_Health_Organization_2009].
-In Sweden, many hospitals have mortality conferences on deceased patient cases, but only one has MoM conferences to better their trauma care on patients who lived [@SweTrau_2023].
-The reason being that this expansion of MoM conferences takes time and resources, which is why all hospitals don't implement this.
+Many hospitals have a trauma registry where they log the patients
+information and timeline of what happened. The extent of the registry
+varies, with HIC having more complete registries. This logging of
+information is important work, as a cornerstone of trauma quality
+improvement programs is multidisciplinary MoM conferences. The MoM
+conference is a meeting where different specialities who work with
+trauma care sit down and discuss deaths and complications in order to
+look for preventable factors. These conferences are performed in many
+hospitals globally, almost everywhere where there are formal medical
+specialty departments and sometimes in smaller hospitals as well
+[@World_Health_Organization_2009]. In Sweden, many hospitals have
+mortality conferences on deceased patient cases, but only one has MoM
+conferences to better their trauma care on patients who lived
+[@SweTrau_2023]. The reason being that this expansion of MoM conferences
+takes time and resources, which is why all hospitals don't implement
+this.
 
-The endpoint of these MoM conferences are opportunities for improvement (OFI).
-At the end of the conference, consensus is reached regarding the existence of an OFI and implementation of corrective action.
-This process is effective, as it has been shown that this review is associated with high-quality trauma care [@Santana_Stelfox_2014].
-Examples of OFI may include lack of resources and management errors.
-One common OFI that this paper will look further into is delayed treatment.
-According to previous studies, among the preventable or possibly preventable deaths in trauma patients, delay in treatment has been identified as a major error contributing to death, found in up to 52.9% of patients in said group [@Teixeira_Inaba_Hadjizacharia_Brown_Salim_Rhee_Browder_Noguchi_Demetriades_2007].
-Delayed treatment has been shown to have adverse effects on patients, showing why it is such an important issue and the need to develop strategies to combat [@Sampalis_Boukas_Lavoie_Nikolis_Frechette_Brown_Fleiszer_Mulder_1995][@Sampalis_Lavoie_Salas_Nikolis_Williams_1994].
-Although it is such a common OFI, the patient level factors associated with delayed treatment are poorly understood.
-There may be several factors that correlate with receiving delayed treatment, but they have yet to be identified.
-Mapping these factors may help in identifying patients that might be at risk for receiving delayed treatment before it happens.
+The endpoint of these MoM conferences are opportunities for improvement
+(OFI). At the end of the conference, consensus is reached regarding the
+existence of an OFI and implementation of corrective action. This
+process is effective, as it has been shown that this review is
+associated with high-quality trauma care [@Santana_Stelfox_2014].
+Examples of OFI may include lack of resources and management errors. One
+common OFI that this paper will look further into is delayed treatment
+<!-- KS: do not use "this paper" in the introduction -->. According to
+previous studies, among the preventable or possibly preventable deaths
+in trauma patients, delay in treatment has been identified as a major
+error contributing to death, found in up to 52.9% of patients in said
+group
+[@Teixeira_Inaba_Hadjizacharia_Brown_Salim_Rhee_Browder_Noguchi_Demetriades_2007].
+Delayed treatment has been shown to have adverse effects on patients,
+showing why it is such an important issue and the need to develop
+strategies to combat
+[@Sampalis_Boukas_Lavoie_Nikolis_Frechette_Brown_Fleiszer_Mulder_1995][@Sampalis_Lavoie_Salas_Nikolis_Williams_1994].
+Although it is such a common OFI, the patient level factors associated
+with delayed treatment are poorly understood. There may be several
+factors that correlate with receiving delayed treatment, but they have
+yet to be identified. Mapping these factors may help in identifying
+patients that might be at risk for receiving delayed treatment before it
+happens.
 
-**Aim**
-
-The aim of this paper is to determine what different patient level factors affect the risk of receiving delayed treatment at a trauma center.
+**Aim** The aim of this paper is to determine what different patient
+level factors affect the risk of receiving delayed treatment at a trauma
+center.
 
@@
 
+<!-- KS: Follow the STROBE order -->
+
 **Study design**
-
-This is a registry based cohort study using data from the trauma registry and trauma care quality database at the Karolinska University Hospital in Solna.
-The trauma registry includes about 14000 patients treated between 2012 and 2022.
-The trauma care quality database is a subset of the trauma registry and includes about 8000 patients selected for review between 2013 and 2022.
-This project will link the two databases and assess how different patient level factors, such as age, sex, mechanism of injury, and injury severity, are associated with delayed treatment using logistic regression.
+<!-- KS: try to avoid "this is a". Good examples are: "we conducted a" etc. -->
+This is a registry based cohort study using data from the trauma
+registry and trauma care quality database at the Karolinska University
+Hospital in Solna. The trauma registry includes about 14000 patients
+treated between 2012 and 2022. The trauma care quality database is a
+subset of the trauma registry and includes about 8000 patients selected
+for review between 2013 and 2022. This project will link the two
+databases and assess how different patient level factors, such as age,
+sex, mechanism of injury, and injury severity, are associated with
+delayed treatment using logistic regression.
 
@@
 
-The outcome is delayed treatment, as identified by the multidisciplinary review board and recorded in the trauma care quality database.
+The outcome is delayed treatment, as identified by the multidisciplinary
+review board and recorded in the trauma care quality database.
 
-**Setting**
-
-Every year, Karolinska University hospital, which is the equivalent of a trauma level 1 hospital, admits 1500 trauma patients each year to their facility.
-To be added to the trauma registry, a patient must be over 15 years of age, had ISS\>9 and/or had trauma code activation.
-They get added to both the Karolinska Trauma registry but also the national trauma registry (SweTrau).
-The registry includes data on vital signs, injuries, interventions and patient information.
-A subset to this database, the trauma care quality database includes information relevant for MoM conferences.
-At Karolinska, patients who die during hospital stay gets their case sent to a MoM everytime.
-However, for the rest of the patients, it is determined by two specialized nurses.
-Several audit filters are set up to flag for patients with possible OFIs.
-Flagged patients are then reviewed by these two nurses who decide if there is a possible OFI, in which case their case is sent to a MoM conference for further review.
+**Setting** <!-- KS: this should be the first paragraph --> Every year,
+Karolinska University hospital, which is the equivalent of a trauma
+level 1 hospital, admits 1500 trauma patients each year. To be added to
+the trauma registry, a patient must be over 15 years of age, had ISS\>9
+and/or had trauma code activation. They get added to both the Karolinska
+Trauma registry as well as the national trauma registry (SweTrau). The
+registry includes data on vital signs, injuries, interventions and
+patient information. A subset to this database
+<!-- KS: the trauma care quality database is not a subset of the trauma registry -->,
+the trauma care quality database includes information relevant for MoM
+conferences. At Karolinska University Hospital, patients who die during
+hospital stay gets their case sent to a MoM everytime. However, for the
+rest of the patients, it is determined by two specialized nurses
+<!-- KS: what is determined? try to avoid "it" and "this" and be specific in scientific texts -->.
+Several audit filters are set up to flag for patients with possible
+OFIs. Flagged patients are then reviewed by these two nurses who decide
+if there is a possible OFI, in which case their case is sent to a MoM
+conference for further review. <!-- KS: add a flowchart/figure -->
 
@@
 
-The database only includes patients 15 and above, ISS \>9 and/or trauma team activation.
-Inclusion in the study further requires that the patients have been assessed for OFIs.
-Patients were excluded if they were missing information in one of the examined patient factors.
+The database only includes patients 15 <!-- KS: -years-old --> and
+above, ISS \>9 and/or trauma team activation. Inclusion in the study
+further requires that the patients have been assessed for OFIs. Patients
+were excluded if they were missing information in one of the examined
+patient factors.
 
@@
 
-The outcome is delayed treatment, which is defined as more than 30 minutes passing from arrival before receiving treatment, as identified by the multidisciplinary review board and recorded in the trauma care quality database.
+The outcome is delayed treatment, which is defined as more than 30
+minutes passing from arrival before receiving treatment, as identified
+by the multidisciplinary review board and recorded in the trauma care
+quality database.
 
@@
 
-The data comes from the trauma registry and trauma care quality database.
+The data comes from the trauma registry and trauma care quality
+database.
 
@@
 
-Selection bias as this is a retrospective cohort study where the outcome has occurred.
+Selection bias as this is a retrospective cohort study where the outcome
+has occurred.
 
-Misclassification bias as the conference may reach a consensus that is wrong.
+Misclassification bias as the conference may reach a consensus that is
+wrong.
 
@@
 
-The study size is all of the eligible patients that was treated at the hospital between 2012 and 2022.
-Starting 2013 only a subgroup of patients were screened for OFIs, but starting 2017 all patients that are included in the database are screened for OFIs.
+The study size is all of the eligible patients that was treated at the
+hospital between 2012 and 2022. Starting 2013 only a subgroup of
+patients were screened for OFIs, but starting 2017 all patients that are
+included in the database are screened for OFIs.
 
@@
 
-Multivariable and potentially univariable logistical regression. A 5% significance level and 95% confidence levels will be used.
+Multivariable and potentially univariable logistical regression. A 5%
+significance level and 95% confidence levels will be used.

### 2023-09-23 manuscriptlongversion.Rmd
intro and method

--- 2023-09-22 manuscriptlongversion.Rmd
@@
 output:
-  word_document: default
+  pdf_document: default
   html_document:
     df_print: paged
-  pdf_document: default
+  word_document: default
 bibliography: bibliography.bib
+editor_options:
+  markdown:
+    wrap: 72
 ---
@@
 
-Trauma, defined as the clinical entity composed of physical injury and the body's associated response, is and has long been a major cause of death around the world today.
-Within modern hospital care, the trauma system stands as an important part in healthcare and is crucial to lowering mortality and morbidity in injured patients through many means, such as hospital care, follow up and prevention programs.
+Trauma, defined as the clinical entity composed of physical injury and
+the body's associated response [@gerdin_risk_2015], is and has long been
+a major cause of death around the world today [@WHO_2021]. Within modern
+hospital care, the trauma system stands as an important part in
+healthcare and is crucial to lowering mortality and morbidity in injured
+patients through many means, such as hospital care, follow up and
+prevention programs [@Choi_2021].
 
-**Trauma statistics**
+## Trauma Statistics
 
-Trauma is often divided into two subgroups, blunt force trauma and penetrating trauma.
-Blunt force trauma is when an object or force strikes the body, often causing bruising, broken bones or deep cuts.
-Examples of blunt trauma could be car crashes, falls or direct blows to the body.
-Penetrating trauma is when an object pierces the skin or body and creates an open wound.
-Examples of penetrating trauma are gunshot wounds and stab wounds [@National_Institute_of_General_Medical_Sciences_2020].
-Differences exist between the two groups, as blunt trauma patients tend to be more injured on arrival to the hospital and as such require more resources and be hospitalized for a longer period because of their more severe injuries [@FITCH20191201].
+Trauma is often divided into two subgroups, blunt force trauma and
+penetrating trauma. Blunt force trauma is when an object or force
+strikes the body, often causing bruising, broken bones or deep cuts.
+Examples of blunt trauma could be car crashes, falls or direct blows to
+the body. Penetrating trauma is when an object pierces the skin or body
+and creates an open wound. Examples of penetrating trauma are gunshot
+wounds and stab wounds [@NIGMS_2020]. Differences exist between the two
+groups, as blunt trauma patients tend to be more injured on arrival to
+the hospital and as such, require more resources and are hospitalized
+for a longer periods [@FITCH2019].
 
-Today, trauma takes the lives of around 4,4 million people each year, almost 8% of all deaths [@World_Health_Organization_2021].
-In the United States, trauma is the 4th leading cause of death among the general population and the leading cause for people between the ages of 1 and 44 [@AAST_2020].
-It is also an important cause of hospitalization and morbidity among all age groups, including seniors, and is responsible for an estimated 10% of all years lived with a disability globally.
-This has a significant burden on social and economic level, costing countries billions of US dollars each year in healthcare and lost productivity[@connolly_woo_lampron_perry_2018].
-Studies estimate the cost of trauma care being between 18,500\$ and 41,500\$ in high income countries (HIC), depending on the country [@Willenberg_Curtis_Taylor_Jan_Glass_Myburgh_2012].
-However, it has been shown that this burden is not evenly distributed between or within countries.
-Many social factors such as age, sex and social status play a major role in the risk of dying from trauma, with young men with low socio-economic status being at most risk [@World_Health_Organization_2021].
-But it is not only patient level factors that affect what effect trauma has on people.
-About 90% of all trauma related deaths occur in low- and middle-income countries (LMIC), with death rates by trauma also being higher than in high income countries.
-Even within these LMIC, people of poorer socio-economic status have higher death rates from trauma.
-Problems identified within these countries that contribute to these statisitics were infrastructure, education and training, attributed to lack of funding, brain drain to HIC and lack of availability of basic amenities [@Shanthakumar_Payne_Leitch_Alfa-Wali_2021].
-In HIC where funding and governance over the healthcare system is better designed, functioning trauma systems and dedicated trauma centers exist that have been shown to lower mortality but also improve functional outcome in trauma patients [@Nirula_Brasel_2006].
+Today, trauma takes the lives of around 4,4 million people each year,
+almost 8% of all deaths [@WHO_2021]. In the United States, trauma is the
+4th leading cause of death among the general population and the leading
+cause for people between the ages of 1 and 44 [@AAST_2020]. It is also
+an important cause of hospitalization and morbidity among all age
+groups, including seniors, and is responsible for an estimated 10% of
+all years lived with a disability globally. This has a significant
+burden on social and economic level, costing countries billions of US
+dollars each year in healthcare and lost productivity [@connolly_2018].
+Studies estimate the cost of trauma care being between 18,500\$ and
+41,500\$ per patient in high income countries (HIC), depending on the
+country [@Willenberg_2012]. However, it has been shown that this burden
+is not evenly distributed between or within countries. Many social
+factors such as age, sex and social status play a major role in the risk
+of dying from trauma, with young men with low socio-economic status
+being at most risk [@WHO_2021]. But it is not only patient level factors
+that affect what effect trauma has on people. About 90% of all trauma
+related deaths occur in low- and middle-income countries (LMIC), with
+death rates by trauma also being higher than in high income countries.
+Even within these LMIC, people of poorer socio-economic status have
+higher death rates from trauma. Problems identified within these
+countries that contribute to these statisitics were infrastructure,
+education and training, attributed to lack of funding, brain drain to
+HIC and lack of availability of basic amenities [@Shanthakumar_2021]. In
+HIC where funding and governance over the healthcare system is better
+designed, functioning trauma systems and dedicated trauma centers exist
+that have been shown to lower mortality but also improve functional
+outcome in trauma patients [@Nirula_2006].
 
-**Trauma systems**
+## Trauma systems
 
-Trauma systems are infrastructures that exist to provide and optimize care for injured patients starting with injury recognition and triage, transport to appropriate trauma center, inpatient care and outpatient follow-up.
-Beyond the clinical side, trauma systems work with outreach, education and advocacy, data collection through registries, research, funding, and disaster preparedness and response.
-A comprehensive and functioning trauma system requires strong leadership and engagement at the trauma center, regional and national level [@connolly_woo_lampron_perry_2018].
-This system is crucial to provide care for trauma patients, both in reducing morbidity and mortality in this patient group.
-Earlier studies from Sweden have shown that treating severely injured patients at a trauma center is associated with a 41% lower adjusted 30-day mortality rate compared to being treated at a non trauma care center due to them being more capable of treating these patients, with potential survival benefit increasing with higher injury severity [@Candefjord_Asker_Caragounis_2020].
-Other studies have shown similar results, showing that treating injured patients at trauma centers is associated with a 15-30% decrease in mortality [@Moran_Lecky_Bouamra_Lawrence_Edwards_Woodford_Willett_Coats_2018][@Celso_Tepas_Langland-Orban_Pracht_Papa_Lottenberg_Flint_2006][@MacKenzie_Rivara_Jurkovich_Nathens_Frey_Egleston_Salkever_Scharfstein_2006][@Haas_Stukel_Gomez_Zagorski_De_Mestral_Sharma_Rubenfeld_Nathens_2012].
-During the last three decades, the introduction of trauma systems has contributed to lowering the incidence of preventable death .
-This is attributed to the improvements in care for acute brain injuries and bleeding control.
-The incidence of late death because of sepsis and multiple organ failure, possibly a result of better and earlier resuscitation [@Asensio_Trunkey_2008].
-Functioning trauma systems saves lives in the hospital, but their work outside of the hospital is just as important.
-Data collection from injured patients, such as mechanism of injury or mortality, are essential for creating databases that can be used for research.
-In turn, that research can be used for planning injury prevention programs that target the most common injuries in the right ways, e.g.
-teen drivers, children, specific occupations etc.
-These injury prevention programs can be planned on trauma center, organizational or government level[@ACS].
+Trauma systems are infrastructures that exist to provide and optimize
+care for injured patients starting with injury recognition and triage,
+transport to appropriate trauma center, inpatient care and outpatient
+follow-up. Beyond the clinical side, trauma systems work with outreach,
+education and advocacy, data collection through registries, research,
+funding, and disaster preparedness and response. A comprehensive and
+functioning trauma system requires strong leadership and engagement at
+the trauma center, regional and national level [@connolly_2018]. This
+system is crucial to provide care for trauma patients, both in reducing
+morbidity and mortality in this patient group. Earlier studies from
+Sweden have shown that treating severely injured patients at a trauma
+center is associated with a 41% lower adjusted 30-day mortality rate
+compared to being treated at a non trauma care center due to them being
+more capable of treating these patients, with potential survival benefit
+increasing with higher injury severity [@Candefjord_2020]. Other studies
+have shown similar results, showing that treating injured patients at
+trauma centers is associated with a 15-30% decrease in mortality
+[@Moran_2018][@Celso_2006][@MacKenzie_2006][@Haas_2012]. During the last
+three decades, the introduction of trauma systems has contributed to
+lowering the incidence of preventable death . This is attributed to the
+improvements in care for acute brain injuries and bleeding control. The
+incidence of late death because of sepsis and multiple organ failure,
+possibly a result of better and earlier resuscitation [@Asensio_2008].
+Functioning trauma systems saves lives in the hospital, but their work
+outside of the hospital is just as important. Data collection from
+injured patients, such as mechanism of injury or mortality, are
+essential for creating databases that can be used for research. In turn,
+that research can be used for planning injury prevention programs that
+target the most common injuries in the right ways, e.g. teen drivers,
+children, specific occupations etc. These injury prevention programs can
+be planned on trauma center, organizational or government level[@ACS].
 
-Trauma teams are multidisciplinary and operate in these centers.
-They play a pivotal role in the treatment of the trauma patient, as they provide the initial care in the critical stage of trauma.
-In Sweden, trauma teams are lead by a team leader who is a surgeon, and include practitioners from the specialities of intensive care, orthopedics, nursing and support staff.
-For the trauma team to be mobilized, trauma code has to be activated, often by an emergency nurse.
-The nurse uses information gathered by the first responders to see if the patient fits any of the criteria for trauma code.
-Most healthcare facilities have established criteria or guidelines that trigger trauma code activation.
-These criteria typically include mechanisms of injury, physiological criteria, anatomical criteria and other specific indicators of severe trauma.
-There are different levels of trauma code with different criteria correlating with severity of injury, with level 1 mobilizing the most personnel to the trauma room.
-Once in the trauma room, the team works systematically to manage the patients injuries.
-They handle the most urgent problems first, such as airways and breathing, with the aim to rapidly assess and stabilise the patient, prioritise their injuries and arrange for site of definitive care.
-[@Hedberg_Falkén_Wihlke_Jansson_Hultgren_Eriksson_Beckman_Haupt_Kjörholt_Sandersjöö_et_al._2020].
+Trauma teams are multidisciplinary and operate in these centers. They
+play a pivotal role in the treatment of the trauma patient, as they
+provide the initial care in the critical stage of trauma. In Sweden,
+trauma teams are lead by a team leader who is a surgeon, and include
+practitioners from the specialities of intensive care, orthopedics,
+nursing and support staff. For the trauma team to be mobilized, trauma
+code has to be activated, often by an emergency nurse. The nurse uses
+information gathered by the first responders to see if the patient fits
+any of the criteria for trauma code. Most healthcare facilities have
+established criteria or guidelines that trigger trauma code activation.
+These criteria typically include mechanisms of injury, physiological
+criteria, anatomical criteria and other specific indicators of severe
+trauma. There are different levels of trauma code with different
+criteria correlating with severity of injury, with level 1 mobilizing
+the most personnel to the trauma room. Once in the trauma room, the team
+works systematically to manage the patients injuries. They handle the
+most urgent problems first, such as airways and breathing, with the aim
+to rapidly assess and stabilise the patient, prioritise their injuries
+and arrange for site of definitive care. [@Hedberg_2020]
 
-**Mortality and morbidity (MoM) conferences**
+## Mortality and morbidity (MoM) conferences
 
-Many hospitals have a trauma registry where they log the patients information and timeline of what happened.
-The extent of the registry varies, with HIC having more complete registries.
-This logging of information is important work, as a cornerstone of trauma quality improvement programs is multidisciplinary MoM conferences.
-The MoM conference is a meeting where different specialities who work with trauma care sit down and discuss deaths and complications in order to look for preventable factors.
-These conferences are performed in many hospitals globally, almost everywhere where there are formal medical specialty departments and sometimes in smaller hospitals as well [@World_Health_Organization_2009].
-In Sweden, many hospitals have mortality conferences on deceased patient cases, but only one has MoM conferences to better their trauma care on patients who lived [@SweTrau_2023].
-The reason being that this expansion of MoM conferences takes time and resources, which is why all hospitals don't implement this.
+Many hospitals have a trauma registry where they log the patients
+information and timeline of what happened. The extent of the registry
+varies, with HIC having more complete registries. This logging of
+information is important work, as a cornerstone of trauma quality
+improvement programs is multidisciplinary MoM conferences. The MoM
+conference is a meeting where different specialities who work with
+trauma care sit down and discuss deaths and complications in order to
+look for preventable factors. These conferences are performed in many
+hospitals globally, almost everywhere where there are formal medical
+specialty departments and sometimes in smaller hospitals as well
+[@WHO_2009]. In Sweden, many hospitals have mortality conferences on
+deceased patient cases, but only one has MoM conferences to better their
+trauma care on patients who lived [@SweTrau_2023]. The reason being that
+this expansion of MoM conferences takes time and resources, which is why
+all hospitals don't implement this.
 
-The endpoint of these MoM conferences are opportunities for improvement (OFI).
-At the end of the conference, consensus is reached regarding the existence of an OFI and implementation of corrective action.
-This process is effective, as it has been shown that this review is associated with high-quality trauma care [@Santana_Stelfox_2014].
-Examples of OFI may include lack of resources and management errors.
-One common OFI that this paper will look further into is delayed treatment.
-According to previous studies, among the preventable or possibly preventable deaths in trauma patients, delay in treatment has been identified as a major error contributing to death, found in up to 52.9% of patients in said group [@Teixeira_Inaba_Hadjizacharia_Brown_Salim_Rhee_Browder_Noguchi_Demetriades_2007].
-Delayed treatment has been shown to have adverse effects on patients, showing why it is such an important issue and the need to develop strategies to combat [@Sampalis_Boukas_Lavoie_Nikolis_Frechette_Brown_Fleiszer_Mulder_1995][@Sampalis_Lavoie_Salas_Nikolis_Williams_1994].
-Although it is such a common OFI, the patient level factors associated with delayed treatment are poorly understood.
-There may be several factors that correlate with receiving delayed treatment, but they have yet to be identified.
-Mapping these factors may help in identifying patients that might be at risk for receiving delayed treatment before it happens.
+The endpoint of these MoM conferences are opportunities for improvement
+(OFI). At the end of the conference, consensus is reached regarding the
+existence of an OFI and implementation of corrective action. This
+process is effective, as it has been shown that this review is
+associated with high-quality trauma care [@Santana_2014]. Examples of
+OFI may include lack of resources and management errors. One common OFI
+is delayed treatment. According to previous studies, among the
+preventable or possibly preventable deaths in trauma patients, delay in
+treatment has been identified as a major error contributing to death,
+found in up to 52.9% of patients in said group [@Teixeira_2007]. Delayed
+treatment has been shown to have adverse effects on patients, showing
+why it is such an important issue and the need to develop strategies to
+combat [@Sampalis_1995][@Sampalis_1994]. Although it is such a common
+OFI, the patient level factors associated with delayed treatment are
+poorly understood. There may be several factors that correlate with
+receiving delayed treatment, but they have yet to be identified. Mapping
+these factors may help in identifying patients that might be at risk for
+receiving delayed treatment before it happens.
 
-**Aim**
+## Aim
 
-The aim of this paper is to determine what different patient level factors affect the risk of receiving delayed treatment at a trauma center.
+The aim of this paper is to determine what different patient level
+factors affect the risk of receiving delayed treatment at a trauma
+center.
 
@@
 
+**Setting**
+
+Every year, Karolinska University hospital, which is the equivalent of a
+trauma level 1 hospital, admits 1500 trauma patients each year. To be
+added to the trauma registry, a patient must be over 15 years of age,
+had ISS\>9 and/or had trauma code activation. They get added to both the
+Karolinska Trauma registry as well as the national trauma registry
+(SweTrau). The registry includes data on vital signs, injuries,
+interventions and patient information. Another database, the trauma care
+quality database includes information relevant for MoM conferences. At
+Karolinska University Hospital, patients who die during hospital stay
+gets their case sent to a MoM everytime. However, for the rest of the
+patients, two specialized nurses determine whether they are. Several
+audit filters are set up to flag for patients with possible OFIs.
+Flagged patients are then reviewed by these two nurses who decide if
+there is a possible OFI, in which case their case is sent to a MoM
+conference for further review.
+
 **Study design**
 
-This is a registry based cohort study using data from the trauma registry and trauma care quality database at the Karolinska University Hospital in Solna.
-The trauma registry includes about 14000 patients treated between 2012 and 2022.
-The trauma care quality database is a subset of the trauma registry and includes about 8000 patients selected for review between 2013 and 2022.
-This project will link the two databases and assess how different patient level factors, such as age, sex, mechanism of injury, and injury severity, are associated with delayed treatment using logistic regression.
+We conducted a registry based cohort study using data from the trauma
+registry and trauma care quality database at the Karolinska University
+Hospital in Solna. The trauma registry includes about 14000 patients
+treated between 2012 and 2022. The trauma care quality database is a
+subset of the trauma registry and includes about 8000 patients selected
+for review between 2013 and 2022. This project will link the two
+databases and assess how different patient level factors, such as age,
+sex, mechanism of injury, and injury severity, are associated with
+delayed treatment using logistic regression.
 
@@
 
-The outcome is delayed treatment, as identified by the multidisciplinary review board and recorded in the trauma care quality database.
-
-**Setting**
-
-Every year, Karolinska University hospital, which is the equivalent of a trauma level 1 hospital, admits 1500 trauma patients each year to their facility.
-To be added to the trauma registry, a patient must be over 15 years of age, had ISS\>9 and/or had trauma code activation.
-They get added to both the Karolinska Trauma registry but also the national trauma registry (SweTrau).
-The registry includes data on vital signs, injuries, interventions and patient information.
-A subset to this database, the trauma care quality database includes information relevant for MoM conferences.
-At Karolinska, patients who die during hospital stay gets their case sent to a MoM everytime.
-However, for the rest of the patients, it is determined by two specialized nurses.
-Several audit filters are set up to flag for patients with possible OFIs.
-Flagged patients are then reviewed by these two nurses who decide if there is a possible OFI, in which case their case is sent to a MoM conference for further review.
+The outcome is delayed treatment, as identified by the multidisciplinary
+review board and recorded in the trauma care quality database.
 
@@
 
-The database only includes patients 15 and above, ISS \>9 and/or trauma team activation.
-Inclusion in the study further requires that the patients have been assessed for OFIs.
-Patients were excluded if they were missing information in one of the examined patient factors.
+The database only includes patients 15 years old and above, ISS \>9
+and/or trauma team activation. Inclusion in the study further requires
+that the patients have been assessed for OFIs. Patients were excluded if
+they were missing information in one of the examined patient factors.
 

### 2023-09-23 flowchart.R

@@ -1,0 +1,221 @@
+################################
+## Data/numbers for flowchart ##
+################################
+## Load functions
+source("functions/functions.R")
+
+## Import data
+datasets <- rofi::import_data()
+
+## Merge data
+combined.datasets <- rofi::merge_data(datasets)
+## SIC(!) Temp fix
+combined.datasets[combined.datasets$tra_id == "92386","tra_DodsfallsanalysGenomford"] <- 1
+## Create OFI column
+combined.datasets$ofi <- create_ofi2(combined.datasets)
+
+#Remove documentation OFI
+combined.datasets$ofi <- ifelse(grepl("Dokumetation|dokumentation|Dokumentation", combined.datasets$Problemomrade_.FMP), "No", combined.datasets$ofi)
+
+dataset.clean.af <- clean_audit_filters(combined.datasets)
+
+## Separate and store cases without known outcome
+missing.outcome <- is.na(dataset.clean.af$ofi)
+n.missing.outcome <- sum(missing.outcome)
+dataset.clean.af <- dataset.clean.af[!missing.outcome, ]
+
+## sep and store cases with age < 15
+age15 <- dataset.clean.af[dataset.clean.af$pt_age_yrs <= 14, ]
+n.under.15 <- nrow(age15)
+
+dataset.comp.age <- dataset.clean.af[dataset.clean.af$pt_age_yrs > 14, ]
+
+## Fix formating and remove wrong values like 999
+clean.dataset <- clean_all_predictors(dataset.comp.age)
+
+n.not.ofi.screened <- sum(is.na(combined.datasets$ofi))
+
+
+df <- subset(clean.dataset, clean.dataset$tra_DodsfallsanalysGenomford == 1)
+n.df <- nrow(df)
+df2 <- subset(clean.dataset, clean.dataset$Fr1.14 == 2 | clean.dataset$Fr1.14 == 3)
+#df3 <- subset(clean.dataset,clean.dataset$tra_id == 77521)
+df2 <- df2[,c("VK_avslutad","tra_DodsfallsanalysGenomford","Problemomrade_.FMP","Fr1.14","tra_id","ofi","Deceased")]
+### Yes and No from mortality conferance
+n.df.ofi <- sum(df$ofi == "Yes", na.rm = TRUE)
+n.df.no.ofi <- sum(df$ofi == "No", na.rm = TRUE)
+
[…]
